### 2024-02-11 paper/paper.qmd
Added analysis_data.csv in order for paper.qmd to compile. Added code for making figures in paper.qmd and compiled it to generate paper.pdf.

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
library(ggplot2)
library(janitor)
library(dplyr)
library(rnaturalearth)
library(sf)
```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

```{r}
#| echo: false
#| warning: false
#| message: false
cleaned_ghs_index_data_2021 <- read_csv("../data/analysis_data/cleaned_GHS_index.csv")

world <- ne_countries(scale = "medium", returnclass = "sf")

world_scores <- merge(world, cleaned_ghs_index_data_2021, by.x = "name", by.y = "country", all.x = TRUE)

world_scores$score_factor <- cut(
  world_scores$overall_score, 
  breaks = c(16, 25, 30, 35, 40, 45, 50, 60, 70, 76), 
  labels = c("16 to 25", "25 to 30", "30 to 35", "35 to 40", "40 to 45", "45 to 50", "50 to 60", "60 to 70", "70 to 76"), 
  include.lowest = TRUE)

ggplot(data = world_scores) +
  geom_sf(aes(fill = score_factor), color = "white") +
  scale_fill_manual(values = c("#990000", "#CC3333", "#FF6666", "#FF9966", "#FFCC99", "#99CCFF", "#6699CC", "#336699", "#003399")) +
  labs(fill = "Global health security score", title = "Global Health Security Score (2021)") +
  theme_minimal()
```

```{r}
#| echo: false
#| warning: false
#| message: false
cleaned_deaths_data <- read_csv("../data/analysis_data/cleaned_all_cause_dth_rates.csv")

cleaned_deaths_data_long <- pivot_longer(cleaned_deaths_data, cols = -country_name, names_to = "Year", values_to = "Value")

cleaned_deaths_data_long[,-1] <- lapply(cleaned_deaths_data_long[,-1], as.numeric)

cleaned_deaths_data_fit <- cleaned_deaths_data_long |>
  filter(Year >= 2010 & Year <= 2019)

ggplot(cleaned_deaths_data_long, aes(x = Year, y = Value, group = country_name, color = country_name)) +
  geom_line(aes(color = country_name)) +
  geom_point(aes(color = country_name)) +
  geom_smooth(data = cleaned_deaths_data_fit, method = "lm", se = FALSE, aes(color = country_name), linetype = "dashed", size = 0.5) +
  geom_smooth(data = cleaned_deaths_data_long, method = "lm", se = FALSE, aes(color = country_name), linetype = "dotted", size = 1) +
  theme_minimal() +
  scale_x_continuous(breaks = 2010:2021) +
  labs(title = "Values Over Years by Country", x = "Year", y = "Value") +
  theme(legend.title = element_blank())
```

```{r}
#| echo: false
#| warning: false
#| message: false

# Model: India
model_india <- lm(Value ~ Year, data = cleaned_deaths_data_fit |> filter(country_name == "India"))
# Predicting 2020 and 2021
coefficients_ind <- coef(model_india)
ind_2020 <- (coefficients_ind["Year"] * 2020) + coefficients_ind["(Intercept)"]
ind_2021 <- (coefficients_ind["Year"] * 2021) + coefficients_ind["(Intercept)"]
# Ratio
ind <- cleaned_deaths_data_long |>
  filter(country_name == "India")
excess_ind_2020 <- ind$Value[11] - ind_2020
excess_ind_2021 <- ind$Value[12] - ind_2021

# Model: Russian Federation
model_rus <- lm(Value ~ Year, data = cleaned_deaths_data_fit |> filter(country_name == "Russian Federation"))
# Predicting 2020 and 2021
coefficients_rus <- coef(model_rus)
rus_2020 <- (coefficients_rus["Year"] * 2020) + coefficients_rus["(Intercept)"]
rus_2021 <- (coefficients_rus["Year"] * 2021) + coefficients_rus["(Intercept)"]
# Ratio
rus <- cleaned_deaths_data_long |>
  filter(country_name == "Russian Federation")
excess_rus_2020 <- rus$Value[11] - rus_2020
excess_rus_2021 <- rus$Value[12] - rus_2021

# Model: United States
model_us <- lm(Value ~ Year, data = cleaned_deaths_data_fit |> filter(country_name == "United States"))
# Predicting 2020 and 2021
coefficients_us <- coef(model_us)
us_2020 <- (coefficients_us["Year"] * 2020) + coefficients_us["(Intercept)"]
us_2021 <- (coefficients_us["Year"] * 2021) + coefficients_us["(Intercept)"]
# Ratio
us <- cleaned_deaths_data_long |>
  filter(country_name == "United States")
excess_us_2020 <- us$Value[11] - us_2020
excess_us_2021 <- us$Value[12] - us_2021

# Model: South Korea
model_sk <- lm(Value ~ Year, data = cleaned_deaths_data_fit |> filter(country_name == "Korea, Rep."))
# Predicting 2020 and 2021
coefficients_sk <- coef(model_sk)
sk_2020 <- (coefficients_sk["Year"] * 2020) + coefficients_sk["(Intercept)"]
sk_2021 <- (coefficients_sk["Year"] * 2021) + coefficients_sk["(Intercept)"]
# Ratio
sk <- cleaned_deaths_data_long |>
  filter(country_name == "Korea, Rep.")
excess_sk_2020 <- sk$Value[11] - sk_2020
excess_sk_2021 <- sk$Value[12] - sk_2021

excess_deaths <- tibble(
  Year = c(2020, 2020, 2020, 2020, 2021, 2021, 2021, 2021),
  country_name = rep(c("India", "Russian Federation", "United States", "Korea, Rep."), times = 2),
  Value = c(excess_ind_2020, excess_rus_2020, excess_us_2020, excess_sk_2020, excess_ind_2021, excess_rus_2021, excess_us_2021, excess_sk_2021) |> round(2)
)

ggplot(excess_deaths, aes(x = country_name, y = Value, fill = as.factor(Year))) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.7) +
  coord_flip() +  # Flip coordinates to make the bar plot horizontal
  theme_minimal() +
  labs(title = "Value by Country and Year", x = "Country", y = "Value") +
  scale_fill_manual(values = c("2021" = "cyan4", "2020" = "cadetblue2"), name = "Year") +
  geom_text(aes(label = Value), position = position_dodge(width = 0.8), color = "black")
```

```{r}
#| echo: false
#| warning: false
#| message: false
combined_data <- read_csv("../data/analysis_data/cleaned_excess_death_ghs.csv")

ggplot(combined_data, aes(x = ghs_index, y = excess_death_rate, label = country)) +
  geom_point(aes(color = ifelse(country %in% c("USA", "Russia"), "Highlight", "Normal"))) +
  scale_color_manual(values = c("Highlight" = "red", "Normal" = "black")) +
  guides(color = FALSE) +
  geom_text(vjust = 1.2, hjust = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#669900") +
  labs(title = "Scatterplot of Excess Death Rate vs GHS Index",
       x = "Global Health Security Index",
       y = "Excess Death Rate") +
  theme_minimal()
```

# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-02-11 paper/paper.qmd
Introduction section written in the paper. llm/usage.txt updated with prompts and chats. references.bib updated with relevate references and bibliography. The paper being replicated has been added to other/literature.

@@
 ---
-title: "My title"
-subtitle: "My subtitle if needed"
+title: "Why Did Russia Fare So Poorly during COVID: An Analysis of why one of the Best Prepared Countries could not control Excess Deaths"
 author: 
-  - First author
-  - Another author
-thanks: "Code and data are available at: LINK."
+  - Aryaman Sharma
+  - Aviral Bhardwaj
+  - Janel Gilani
+thanks: "Code and data are available at: https://github.com/Ary4m3n/covid-effect-russia.git; Replication on Social Science Reproduction platform available at: "
 date: today
@@
 format: pdf
+toc: true
 number-sections: true
@@
 library(sf)
+library(knitr)
 ```
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+The COVID-19 pandemic has led to an enormous loss of life over the world and has presented an unprecedented challenge to public health, foreign policy and has additionally had an adverse effect on countries' economies around the world, over the past 5 years [@citeWHOeffect].
+There have been over 774 million reported confirmed cases of the virus being contract and the outbreak has caused over 7 million deaths, which are claimed to be "under counted" and "under reported", suggesting that the actual number is even higher around 18.2 - 33.5 million based on the excess mortality estimates [@citeWikiCOVID].
+This excessively high number of deaths has alerted the world to assess how countries assumed to be well-prepared to tackle such outbreaks were affected to this extent.
 
-The remainder of this paper is structured as follows. @sec-data....
+There were a subset of countries, namely India, the United States of America, Brazil, Russia and Mexico which were impacted the most in terms of having the highest cumulative total deaths [@citeThinkGlobalHealth].
+According to the global data, Russia's COVID-19 related deaths since the start of the pandemic have been over 230,000.
+Russia also holds the highest mortality rate in Europe and the second-highest rate in Asia [@citeCarnegieRussia].
+For these reasons and more, Russia and its residents have suffered a lot over the past few years, wherein the adverse impact of the outbreak has lead to a plethora of impacts on the country's economy and standard of living.
+
+This paper will follow a reproduction of Jennifer B.
+Nuzzo and Jorge R.
+Ledesma's paper [@citeRepPaper] and findings and apply a Russian-facing lens to discuss the impact of the COVID-19 outbreak on Russia and what could have been done better and can be improved for the future in order for the impact of any upcoming pandemics can be lessened.
+Our paper aims to replicate their claim of how well-prepared countries did so poorly during the COVID-19 outbreak, where the Nuzzo and Ledesma have concentrated solely on the United States but in this paper we look at Russia with the same regards and analyse the impact of the pandemic on the death toll, and finally provide for ways to improve the laws, regulations and plans to tackle such outbreaks in the future.
+Our reproduction found out that how ever well prepared Russia seemed to be, it experienced an unprecedented rise in the death rate compared to other countries, and its excess death rate was one of the highest amongst all countries.
+This leads us to make an important judgement that better planning and regulations can help reduce the impact of such outbreaks in the future, which will be studied in this paper.
+
+This paper is structured using the following sections: Data, Results and Discussion.
+In the Data (@sec-data) section, the data source of the datasets from the paper being replicated [@citeRepPaper] is discussed and the measurement and data cleaning process is outlined.
+In the Results (@sec-results) section, the paper summarizes the data findings and presents relevant maps and plots in order to study the trends.
+The paper ends with the Discussion (@sec-discussion) section, where the findings of the paper have been analysed and delved deeper into by studying the measures that can be taken to prevent the country from experiencing such an extensive effect of future pandemics.
+Additionally, the limitations and a further scope for the paper has been discussed here.
 
@@
 
-# Results
+# Results {#sec-results}
 
@@
 
-# Discussion
+# Discussion {#sec-discussion}
 

### 2024-02-12 paper/paper.qmd
Minor update to references.bib. Added the Data section to paper.qmd and compiled it. NOTE: Measurement subsection left to do in Data.

@@
 ---
-title: "Why Did Russia Fare So Poorly during COVID: An Analysis of why one of the Best Prepared Countries could not control Excess Deaths"
+title: "Why Did Russia Fare So Poorly during COVID: An Analysis of Why One of the Best Prepared Countries Could Not Control Excess Deaths"
 author: 
@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Aenean tempus libero sit amet nisl dapibus, varius dictum magna pellentesque.
+In lacinia nec ligula eget condimentum.
+Aenean imperdiet enim sit amet felis volutpat, eu ornare ante pretium.
+Donec ullamcorper tempus facilisis.
+Mauris porttitor libero ultrices tortor facilisis, quis varius sapien ultricies.
+Pellentesque et metus vel quam mollis dignissim ultricies a dui.
+Cras sodales porttitor pulvinar.
+Proin vel nunc sit amet sapien venenatis pellentesque nec at felis."
 format: pdf
@@
 Additionally, the limitations and a further scope for the paper has been discussed here.
+
+# Data {#sec-data}
+
+As mentioned above, the data analysed in this paper is from the reproduction package of the paper by Nuzzo and Ledesma [@citeRepPaper].
+The data was cleaned and analysed using the open source `R` programming language [@citeR].
+R libraries and packages such as `tidyverse` [@citeTidyverse], `janitor` [@citeJanitor], `ggplot2` [@citeGgplot], `knitr` [@citeKnitr], `readr` [@citeReadr], `dplyr` [@citeDplyr], `rnaturalearth` [@citeRnaturalearth] and `sf` [@citeSf].
+In the following sections, we will discuss the raw data in (@sec-data-raw), the data cleaning process in (@sec-data-clean) and then move on to discussing the measurement in (@sec-data-measurement).
+
+## Raw Data {#sec-data-raw}
+
+This paper will replicate the data that was obtained from the replication package of the paper [@citeRepPaper].
+Specifically, in the paper we use 3 datasets that will help us analyse the claim of Russia not being able to control the wave of the COVID-19 pandemic well.
+In general, there is a lot of data available on the effect of COVID-19 on countries, however, all the data we needed for this paper was included in the paper by Nuzzo and Ledesma.
+
+The first raw dataset includes data on the Global Health Security index for all countries.
+The GHS index is an indicator of how well prepared or "secure" a country is against epidemics or pandemics like the COVID-19 outbreak.
+This dataset has a plethora of data on indices that are not relevant to us for this paper.
+Additionally, the dataset contains data for years 2019 and 2021.
+For our analysis, we only require data for the GHS index of countries for 2021.
+
+The second raw dataset includes data on all cause death rates for all countries from 1960-2021.
+This dataset has loads of information that was mainly unnecessary to us for this paper, mainly the data from 1960-2010 which is not required as we are only interested in looking at the effect of COVID-19 which is mainly data from 2020-2021.
+However, in order to study the trend, we will use data for all cause death rates for countries around the world from 2010-2021.
+
+The third raw dataset is from the Institute of Health Metrics and Evaluation (IHME) which presents data on the COVID death rate, excess deaths, excess death rate and ratio of excess death rate over COVID death rate for all countries in the world.
+Again, this dataset too contained unnecessary information that was of no need to us which the cleaned data process outlined in @sec-data-clean.
+
+In the next section (@sec-data-clean), we will outline the data-cleaning process and also show the structure of the cleaned data.
+
+## Cleaned Data {#sec-data-clean}
+
+As stated above in @sec-data-raw, we start with the first dataset on GHS indices for all countries in the world.
+@tbl-cleaned-ghs shows the first 6 countries and their GHS indices.
+Here we only have 2 variables, namely the Country Name and the respective GHS index.
+We cleaned this dataset by selecting only the two columns for the respective variables and filtering the data for the year of 2021.
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: tbl-cleaned-ghs
+#| tbl-cap: "Cleaned Data showing GHS indices for first 6 countries"
+#| tbl-cap-location: bottom
+#| warning: false
+
+cleaned_ghs_index_data_2021 <- read_csv("../data/analysis_data/cleaned_GHS_index.csv", show_col_types = FALSE)
+
+table_cleaned_ghs_index_data_2021 <- cleaned_ghs_index_data_2021 |>
+  rename(
+    `Country Name` = country,
+    `GHS Index` = overall_score
+  )
+
+kable(head(table_cleaned_ghs_index_data_2021))
+```
+
+Second, we have a raw dataset on all cause death rates for all countries from 1960-2021.
+We cleaned this dataset by selecting a subset of countries that helped us analyse how Russia fared against other countries in death ratesover the years.
+@tbl-cleaned-death-rates shows the structure of the cleaned data after we filtered it for the time period of 2010-2021 for India, South Korea, Russia and the United States.
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: tbl-cleaned-death-rates
+#| tbl-cap: "Cleaned Data showing all cause death rates for 4 countries"
+#| tbl-cap-location: bottom
+#| warning: false
+
+cleaned_deaths_data <- read_csv("../data/analysis_data/cleaned_all_cause_dth_rates.csv")
+
+cleaned_deaths_data$country_name[cleaned_deaths_data$country_name == "Russian Federation"] <- "Russia"
+cleaned_deaths_data$country_name[cleaned_deaths_data$country_name == "Korea, Rep."] <- "South Korea"
+
+table_cleaned_deaths_data <- cleaned_deaths_data |>
+  rename(
+    `Country Name` = country_name,
+  )
+
+kable(head(table_cleaned_deaths_data))
+```
+
+The final raw dataset, as mentioned in @sec-data-raw, that we have is the IHME data on the COVID death rate, excess deaths, excess death rate and ratio of excess death rate over COVID death rate for all countries in the world.
+In this paper we only needed the two variables excess death rate and the country name, however for better analysis of Russia's ability to control the pandemic's effect, we combined this data with the respective GHS index as shown in @tbl-cleaned-combined.
+Hence, there are three variables, the country name, and the respective excess death rate and GHS index.
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: tbl-cleaned-combined
+#| tbl-cap: "Cleaned Data showing Excess Death Rate and GHS index for first 6 chosen countries"
+#| tbl-cap-location: bottom
+#| warning: false
+
+combined_data <- read_csv("../data/analysis_data/cleaned_excess_death_ghs.csv")
+
+table_combined_data <- combined_data |>
+  rename(
+    `Country Name` = country,
+    `Excess Death Rate` = excess_death_rate,
+    `GHS Index` = ghs_index
+  )
+
+kable(head(table_combined_data))
+```
+
+Now that we have cleaned our raw datasets and have generated clean tables for analysis, we will go on to the Results Section (@sec-results) and delve deeper into studying the effect of the COVID-19 pandemic on Russia and if it was impacted significantly more than other countries.
+
+## Measurement {#sec-data-measurement}
+
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Aenean tempus libero sit amet nisl dapibus, varius dictum magna pellentesque.
+In lacinia nec ligula eget condimentum.
+Aenean imperdiet enim sit amet felis volutpat, eu ornare ante pretium.
+Donec ullamcorper tempus facilisis.
+Mauris porttitor libero ultrices tortor facilisis, quis varius sapien ultricies.
+Pellentesque et metus vel quam mollis dignissim ultricies a dui.
+Cras sodales porttitor pulvinar.
+Proin vel nunc sit amet sapien venenatis pellentesque nec at felis.
+
+# Results {#sec-results}
 
@@
 
-# Data {#sec-data}
-
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
-
-```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
-#| echo: false
-
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
-  theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
-  coord_flip()
-```
-
-Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
-```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
-#| echo: false
-#| warning: false
-#| message: false
-
-analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
-
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
-  theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
-```
-
-Talk way more about it.
-
-# Model
-
-The goal of our modelling strategy is twofold. Firstly,...
-
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
-
-## Model set-up
-
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
-
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
-
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
-
-### Model justification
-
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
-
-# Results {#sec-results}
-
-Our results are summarized in @tbl-modelresults.
-
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-
-library(rstanarm)
-
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
-```
-
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
-
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
-```
-
 # Discussion {#sec-discussion}
@@
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
+<!-- In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows... -->
 
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
+<!-- In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...  -->
 
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
+<!-- ```{r} -->
+<!-- #| eval: true -->
+<!-- #| echo: false -->
+<!-- #| message: false -->
+<!-- #| warning: false -->
+<!-- #| label: fig-ppcheckandposteriorvsprior -->
+<!-- #| layout-ncol: 2 -->
+<!-- #| fig-cap: "Examining how the model fits, and is affected by, the data" -->
+<!-- #| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"] -->
 
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
+<!-- pp_check(first_model) + -->
+<!--   theme_classic() + -->
+<!--   theme(legend.position = "bottom") -->
 
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
-```
+<!-- posterior_vs_prior(first_model) + -->
+<!--   theme_minimal() + -->
+<!--   scale_color_brewer(palette = "Set1") + -->
+<!--   theme(legend.position = "bottom") + -->
+<!--   coord_flip() -->
+<!-- ``` -->
 
-## Diagnostics
+<!-- ## Diagnostics -->
 
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
+<!-- @fig-stanareyouokay-1 is a trace plot. It shows... This suggests... -->
 
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
+<!-- @fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests... -->
 
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
+<!-- ```{r} -->
+<!-- #| echo: false -->
+<!-- #| eval: true -->
+<!-- #| message: false -->
+<!-- #| warning: false -->
+<!-- #| label: fig-stanareyouokay -->
+<!-- #| fig-cap: "Checking the convergence of the MCMC algorithm" -->
+<!-- #| fig-subcap: ["Trace plot", "Rhat plot"] -->
+<!-- #| layout-ncol: 2 -->
 
-plot(first_model, "trace")
+<!-- plot(first_model, "trace") -->
 
-plot(first_model, "rhat")
-```
+<!-- plot(first_model, "rhat") -->
+<!-- ``` -->
+
+<!-- # Data {#sec-data} -->
+
+<!-- Some of our data is of penguins (@fig-bills), from @palmerpenguins. -->
+
+<!-- ```{r} -->
+<!-- #| label: fig-bills -->
+<!-- #| fig-cap: Bills of penguins -->
+<!-- #| echo: false -->
+
+<!-- ggplot(penguins, aes(x = island, fill = species)) + -->
+<!--   geom_bar(alpha = 0.8) + -->
+<!--   scale_fill_manual(values = c("darkorange","purple","cyan4"), -->
+<!--                     guide = "none") + -->
+<!--   theme_minimal() + -->
+<!--   facet_wrap(~species, ncol = 1) + -->
+<!--   coord_flip() -->
+<!-- ``` -->
+
+<!-- Talk more about it. -->
+
+<!-- And also planes (@fig-planes).
+(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.) -->
+
+<!-- ```{r} -->
+<!-- #| label: fig-planes -->
+<!-- #| fig-cap: Relationship between wing length and width -->
+<!-- #| echo: false -->
+<!-- #| warning: false -->
+<!-- #| message: false -->
+
+<!-- analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv")) -->
+
+<!-- analysis_data |>  -->
+<!--   ggplot(aes(x = width, y = length)) + -->
+<!--   geom_point(alpha = 0.8) + -->
+<!--   theme_minimal() + -->
+<!--   labs(x = "Wing width (mm)", -->
+<!--        y = "Wing length (mm)") -->
+<!-- ``` -->
+
+<!-- Talk way more about it.  -->
+
+<!-- # Results {#sec-results} -->
+
+<!-- Our results are summarized in @tbl-modelresults. -->
+
+<!-- ```{r} -->
+<!-- #| echo: false -->
+<!-- #| eval: true -->
+<!-- #| warning: false -->
+<!-- #| message: false -->
+
+<!-- library(rstanarm) -->
+
+<!-- first_model <- -->
+<!--   readRDS(file = here::here("models/first_model.rds")) -->
+<!-- ``` -->
+
+<!-- ```{r} -->
+<!-- #| echo: false -->
+<!-- #| eval: true -->
+<!-- #| label: tbl-modelresults -->
+<!-- #| tbl-cap: "Explanatory models of flight time based on wing width and wing length" -->
+<!-- #| warning: false -->
+
+<!-- modelsummary::modelsummary( -->
+<!--   list( -->
+<!--     "First model" = first_model -->
+<!--   ), -->
+<!--   statistic = "mad", -->
+<!--   fmt = 2 -->
+<!-- ) -->
+<!-- ``` -->
 

### 2024-02-12 paper/paper.qmd
Results section completed and compiled with additions to Appendix too. usage.txt updated with one more prompt that was used. Discussion section left.

@@
 
-# Introduction
+# Introduction {#sec-introduction}
 
@@
 
-cleaned_deaths_data <- read_csv("../data/analysis_data/cleaned_all_cause_dth_rates.csv")
+cleaned_deaths_data <- read_csv("../data/analysis_data/cleaned_all_cause_dth_rates.csv", show_col_types = FALSE)
 
@@
 
-combined_data <- read_csv("../data/analysis_data/cleaned_excess_death_ghs.csv")
+combined_data <- read_csv("../data/analysis_data/cleaned_excess_death_ghs.csv", show_col_types = FALSE)
 
@@
 
+This section will be divided into three parts where all together we aim to look at Russia's capabilty according to the GHS index to tackle pandemics (@sec-results-ghs), then compare Russia's trend in all cause death rates and excess death rates with those of other countries like India, South Korea and the United states (@sec-results-death-rates) and finally study the relationship between the excess death rate and GHS index for Russia along with a subset of countries in the world (@sec-results-death-rates).
+
+## Global Distribution of Global Health Security Index (GHS) in 2021 {#sec-results-ghs}
+
+As explained in @sec-data-raw, the GHS index is an indicator of how well prepared a country is to tackle epidemics or pandemics like the COVID-19 outbreak.
+The GHS index gives one an idea of how the sound health care system  as well as the rules and regulations of a certain country are in the context of facing pandemics.
+Generally, the higher the GHS index of a country, the more well-prepared and secure they are.
+@fig-map shows a representation of the world map with countries shaded on the basis of their GHS indices.
+This map is a replication of figure 1 in Nuzzo and Ledesma paper [@citeRepPaper] that is being replicated in this report.
+
 ```{r}
+#| label: fig-map
+#| fig-cap: Global Distribution of Global Health Security (GHS) Index Scores in 2021
 #| echo: false
@@
 #| message: false
-cleaned_ghs_index_data_2021 <- read_csv("../data/analysis_data/cleaned_GHS_index.csv")
+#| fig-height: 6
+#| fig-width: 8
+cleaned_ghs_index_data_2021 <- read_csv("../data/analysis_data/cleaned_GHS_index.csv", show_col_types = FALSE)
 
@@
   scale_fill_manual(values = c("#990000", "#CC3333", "#FF6666", "#FF9966", "#FFCC99", "#99CCFF", "#6699CC", "#336699", "#003399")) +
-  labs(fill = "Global health security score", title = "Global Health Security Score (2021)") +
+  labs(fill = "GHS Indices") +
   theme_minimal()
@@
 
+We can observe here that there are countries like the United States and Australia whose GHS indices are one of the highest, implying that they were deemed extremely well prepared for epidemics.
+Then we see countries like Russia and China who were, although moderately when compared to the aforementioned countries, but still well prepared to tackle pandemics.
+Finally, we see countries such as Venezuela and Yemen which were said to be the least secure of all.
+
+@fig-map provides this report with a background on how secure Russia was in tackling epidemics and pandemics, which as seen in the map, it was in a good position to face such outbreaks where numerous countries were deemed less prepared than Russia.
+This will aid us in drawing meaningful analyses throughout the paper, specifically in @sec-results-death-rates, sec-results-death-rates and @sec-discussion.
+
+## Trend in Death Rates for Russia, India, South Korea and United States {#sec-results-death-rates}
+
+Now we move on to analyzing the trends in the all cause death rates per 1000 population over 2010-2021, which includes COVID deaths as well as general deaths which were not caused by the outbreak, and the trend in the excess death rates per 100 population in 2020 and 2021.
+The excess death rates in this section were calculated by finding a difference between the expected death rate and the actual death rate for that year.
+
+@fig-all-cause-death-rate shows us the trend for the all cause death rates per 1000 population over 2010-2021 for India, South Korea, Russia and the United States using the data presented in @tbl-cleaned-death-rates.
+We observe here that from the years 2010-2019, there is a general gradually decreasing trend in the death rates in the case of Russia and India, and a gradually increasing trend in the death rates in the case of South Korea and the United States.
+The dashed line represents the line of best fit for the trend from 2010-2019.
+
 ```{r}
+#| label: fig-all-cause-death-rate
+#| fig-cap: All Cause Death Rates per 1000 Population for India, South Korea, Russia and United States (2010-2021)
 #| echo: false
@@
 #| message: false
-cleaned_deaths_data <- read_csv("../data/analysis_data/cleaned_all_cause_dth_rates.csv")
+#| fig-height: 4
+#| fig-width: 8
+cleaned_deaths_data <- read_csv("../data/analysis_data/cleaned_all_cause_dth_rates.csv", show_col_types = FALSE)
+
+cleaned_deaths_data$country_name[cleaned_deaths_data$country_name == "Russian Federation"] <- "Russia"
+cleaned_deaths_data$country_name[cleaned_deaths_data$country_name == "Korea, Rep."] <- "South Korea"
 
@@
   scale_x_continuous(breaks = 2010:2021) +
-  labs(title = "Values Over Years by Country", x = "Year", y = "Value") +
+  labs(title = "All Cause Death Rates by Country (2010-2021)", x = "Year", y = "All Cause Death Rates") +
   theme(legend.title = element_blank())
@@
 
+We see here in @fig-all-cause-death-rate that there is a sudden increase in the death rate in 2020 and 2021, and the increase in Russia's death rate is much higher than that of the other countries.
+The essential point to note here is that in the case of India, which has a GHS index lower than that of Russia, the rising trend observed in 2020 and 2021 was still not as excessive as in the case of Russia.
+The dotted line represents the line of best fit for the trend from 2010-2021.
+As we go from comparing Russia to South Korea, we see that in the case of South Korea, both the dashed (2010-2019) and the dotted (2010-2021) are very close together implying that the country did not experience an unanticipated number of deaths during COVID.
+However, in the case of Russia, we see the two lines diverging implying that there were unforeseen deaths during that period.
+
+This observation leads us to @fig-excess-death-rate, where we present the excess death rate for the four countries for 2020 and 2021.
+As mentioned above, the excess death rates were calculated by extrapolating the dashed line of best-fit from years 2010-2019 in @fig-all-cause-death-rate representing the trend in all cause death rates to 2020 and 2021.
+This way we were able to find the expected death rate for each of the four countries and subsequently we were able to plot the difference between the expected death rate and the actual death rate called the excess death rate for 2020 and 2021 for all four countries in @fig-excess-death-rate.
+
 ```{r}
+#| label: fig-excess-death-rate
+#| fig-cap: Excess Death Rates per 1000 Population for India, South Korea, Russia and United States, 2020 and 2021
 #| echo: false
@@
 #| message: false
+#| fig-height: 4
+#| fig-width: 8
 
@@
 
-# Model: Russian Federation
-model_rus <- lm(Value ~ Year, data = cleaned_deaths_data_fit |> filter(country_name == "Russian Federation"))
+# Model: Russia
+model_rus <- lm(Value ~ Year, data = cleaned_deaths_data_fit |> filter(country_name == "Russia"))
 # Predicting 2020 and 2021
@@
 rus <- cleaned_deaths_data_long |>
-  filter(country_name == "Russian Federation")
+  filter(country_name == "Russia")
 excess_rus_2020 <- rus$Value[11] - rus_2020
@@
 # Model: South Korea
-model_sk <- lm(Value ~ Year, data = cleaned_deaths_data_fit |> filter(country_name == "Korea, Rep."))
+model_sk <- lm(Value ~ Year, data = cleaned_deaths_data_fit |> filter(country_name == "South Korea"))
 # Predicting 2020 and 2021
@@
 sk <- cleaned_deaths_data_long |>
-  filter(country_name == "Korea, Rep.")
+  filter(country_name == "South Korea")
 excess_sk_2020 <- sk$Value[11] - sk_2020
@@
   Year = c(2020, 2020, 2020, 2020, 2021, 2021, 2021, 2021),
-  country_name = rep(c("India", "Russian Federation", "United States", "Korea, Rep."), times = 2),
+  country_name = rep(c("India", "Russia", "United States", "South Korea"), times = 2),
   Value = c(excess_ind_2020, excess_rus_2020, excess_us_2020, excess_sk_2020, excess_ind_2021, excess_rus_2021, excess_us_2021, excess_sk_2021) |> round(2)
@@
   theme_minimal() +
-  labs(title = "Value by Country and Year", x = "Country", y = "Value") +
+  labs(title = "Excess Death Rate by Country and Year", x = "Country", y = "Excess Death Rate") +
   scale_fill_manual(values = c("2021" = "cyan4", "2020" = "cadetblue2"), name = "Year") +
-  geom_text(aes(label = Value), position = position_dodge(width = 0.8), color = "black")
+  geom_text(aes(label = Value), position = position_dodge(width = 0.8), hjust = -0.01, color = "black")
 ```
 
+We see in @fig-excess-death-rate the massive difference between the excess death rates for Russia and the other countries.
+Comparing Russia to South Korea we see that, as seen in @fig-all-cause-death-rate, South Korea was very faintly affected by the COVID-19 pandemic, however Russia saw a sudden rise in deaths from 2019 to 2021.
+This shows us that the observations here are analogous to the plot in @fig-all-cause-death-rate.
+When we compare Russia with India, which as aforementioned was deemed a less secure country in the case of epidemics, we see that the excess death rates for Russia in both 2020 and 2021 are much higher than those for India.
+This observation will further aid us in making meaningful discussions in @sec-discussion.
+
+## Relationship between the Excess Death Rate and Global Health Security Index (GHS) {#sec-results-death-rates}
+
+Finally, we will now explore the relationship between the excess death rate seen above and the GHS index for a subset of countries per 100,000 population.
+@fig-excess-death-rate-ghs shows us a subset of countries scattered around the plot according to their excess death rate and the GHS index shown in the Appendix @tbl-appendix.
+The green line shows the general decreasing trend which implies that as the GHS index increases, the excess death rate tends to decrease, i.e.
+the excess death rate for more "secure" or well-prepared countries is generally lower.
+
+Russia, labelled in red, is seen to be significantly higher than the line of best fit, which implies that Russia's excess death rate was much worse than it should have been.
+We observe that countries like Brazil and Indonesia, which have approximately the same GHS index as Russia, are nearer to the line of best fit than Russia.
+India, although having a lower GHS index than Russia is seen to have a much lower than general excess death rate.
+
 ```{r}
+#| label: fig-excess-death-rate-ghs
+#| fig-cap: Relationship between the Excess Death Rate and the Global Health Security Index (GHS) per 100,000 Population for a subset of countries, 2020-2021
 #| echo: false
@@
 #| message: false
-combined_data <- read_csv("../data/analysis_data/cleaned_excess_death_ghs.csv")
+#| fig-height: 5
+#| fig-width: 8
+
+combined_data <- read_csv("../data/analysis_data/cleaned_excess_death_ghs.csv", show_col_types = FALSE)
 
 ggplot(combined_data, aes(x = ghs_index, y = excess_death_rate, label = country)) +
-  geom_point(aes(color = ifelse(country %in% c("USA", "Russia"), "Highlight", "Normal"))) +
+  geom_point(aes(color = ifelse(country %in% c("Russia"), "Highlight", "Normal"))) +
   scale_color_manual(values = c("Highlight" = "red", "Normal" = "black")) +
   guides(color = FALSE) +
-  geom_text(vjust = 1.2, hjust = 0.5) +
+  geom_text(vjust = 1.5, hjust = 0.5) +
   geom_smooth(method = "lm", se = FALSE, color = "#669900") +
-  labs(title = "Scatterplot of Excess Death Rate vs GHS Index",
-       x = "Global Health Security Index",
+  labs(title = "Excess Death Rate vs GHS Index for a subset of countries",
+       x = "Global Health Security Index (GHS)",
        y = "Excess Death Rate") +
@@
 ```
+
+The implications from this figure will be further studied in @sec-discussion.
 
@@
 
-## Weaknesses and next steps
+## Weaknesses and Further Scope {#sec-disc-weaknesses}
 
@@
 
-# Model details {#sec-model-details}
+```{r}
+#| echo: false
+#| eval: true
+#| label: tbl-appendix
+#| tbl-cap: "Cleaned Data showing Excess Death Rate and GHS index for a subset of countries"
+#| tbl-cap-location: bottom
+#| warning: false
 
-## Posterior predictive check
+combined_data <- read_csv("../data/analysis_data/cleaned_excess_death_ghs.csv", show_col_types = FALSE)
+
+table_combined_data <- combined_data |>
+  rename(
+    `Country Name` = country,
+    `Excess Death Rate` = excess_death_rate,
+    `GHS Index` = ghs_index
+  )
+
+kable(table_combined_data)
+```
 

### 2024-02-13 paper/paper.qmd
Added Discussion section and compiled. Updated references.bib.

@@
 
-## First discussion point {#sec-first-point}
+## Findings: Russia, Excess Deaths and what went wrong during the COVID-19 pandemic {#sec-first-point}
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+In this paper, we have replicated the paper by Nuzzo and Ledesma [@citeRepPaper], where they focused their analysis on the United States and studying how the world's best prepared country still couldn't control the pandemic.
+Our paper has instead shifted the focus to another global power and fairly well prepared country Russia, where we found out that Russia, when compared to a subset of other countries:(1) had one of the highest increase in all cause death rate in 2020, and (2) had one of the highest excess death rate caused by the pandemic in 2020 and 2021, as observed in @fig-all-cause-death-rate and @fig-excess-death-rate.
+Our paper also found and observed the relationship between the GHS index of a country and its excess death rate in @fig-excess-death-rate-ghs, where Russia was seen as an outlier, where it lied a lot higher than the line of best fit implying a higher excess death rate than the other countries around it, even with a similar GHS index.
 
-## Second discussion point
+The main question that this paper tackles is, what exactly did Russia do wrong during the pandemic considering it was the first country to roll out a vaccine?
+According to a report by Carnegie Endowment for International Peace, by late October 2021, only one-third of the Russian Population had been fully vaccinated against the virus, where other countries like Argentina and China had 56% and 76% of their populations respectively, fully vaccinated by then [@citeCarnegieRussia].
+According to reports, one of the main concerns for people being skeptical about the Sputnik V vaccine was that they did not trust that the vaccine would be effective or even safe to use.
+Another reason for the residents of Russia being skeptical was that Russian President Putin hesitated for months after the release of the vaccine to himself get vaccinated [@citeCarnegieRussia].
+I believe that this must have slowed down the rate at which people in Russia were getting vaccinated hence leading to unprecedented deaths as a result.
 
-## Third discussion point
+Another possible reason for Russia not being able to tackle the pandemic well was that the quarantine measures during the outbreak in 2020 and 2021 were generally very sporadic and short [@citeNewYorker].
+It was reported that at a time when the cases testing positive reached nearly ten thousand a day, the government put lockdown restrictions but that too only for a week, before everything opened back up.
+I believe this would have been a huge contributing factor to the extremely high excess deaths as more people would have contracted the virus coming in contact with each, hence spreading it more.
+Additionally, keeping in mind that at this time, as aforementioned, Russia only had approximately one thirds of its population vaccinated, this would have hit the residents of the country really hard.
+
+In @sec-second-point, we will discuss the measures that Russia could have taken to tackle the pandemic in a better way, and also measures that can be taken for future pandemics.
+
+## Russia and Improvements for the future {#sec-second-point}
+
+I believe that, as fast as Russia was at releasing the vaccine, maybe the government could have focused on mass producing and advertising the vaccine as a necessity as well.
+I believe that as soon as there is any skepticism in regards with vaccines, people will choose not to get one.
+As mentioned above, this will lead to the virus spreading even faster and eventually to more deaths too.
+In order to prevent this, the government should show the residents of the country that the vaccine is trustworthy and effective.
+It is essential to note that in order for this to work, the vaccine's efficacy and safety will also have to be proven to prevent from any ethical implications rising.
+
+It was claimed that at the beginning of the outbreak, Russia was trying to downplay the threat of virus, however, with time they realized how lethal it actually was [@citeCarnegieRussia].
+According to me, it is very essential for a country to recognize the potential threat from any epidemic or outbreak and then make laws, and rules and regulations around it to quickly contain the situation before it gets worse.
+For countries as large as Russia, it might be a particularly challenging task to get such a large population that is spread out under control, but if the right procedures are set in place to tackle such situations, it can be manageable.
+Additionally, it is essential in these situations to also have firm rules in place for quarantining and lockdowns to prevent the virus from spreading.
+This is especially even more important in the early stages of the outbreak when there is no vaccine in place to prevent the citizens of a country from contracting the virus.
+Some countries during the COVID-19 pandemic were very strict with their restrictions and lockdowns, and hence experienced way lesser deaths when compared to other countries which were more relaxed.
+In the future too, for Russia to be able to control the situation early on, better and more strict regulations will be required.
+
+## Accounting for Bias and Ethical Implications {#sec-third-point}
+
+It is essential to take into account an unintended but highly likely bias that might have taken place in this paper.
+While working on the paper, we realized that in creating @fig-all-cause-death-rate, @fig-excess-death-rate and @fig-excess-death-rate-ghs, selection bias might have been involved in make decisions on which countries to plot on the graphs.
+Specifically, we used 4 countries in @fig-all-cause-death-rate and @fig-excess-death-rate to show the unique trends for Russia during the pandemic.
+Maybe if we chose a set of different countries, some countries would have performed more like Russia and consequently would have impacted our analysis.
+Similarly, in @fig-excess-death-rate-ghs choosing a different subset of countries might generate a completely different line of best fit which might not have satisfied our exploration in the paper.
+However, at the same time, it is essential to note that the set of countries chosen in this paper has mostly been random and some countries like India and the United States have been shown constantly to create parallels between the impact of the pandemic on both the countries.
+
+There are a few ethical implications that we found while working on the paper.
+Mainly, sometimes the data can be incomplete or hidden which is misleading while analyzing that data.
+It is essential to keep that into account when looking at a report as, if for in this case any data for the deaths were incomplete, that would directly impact how we deem Russia to have tackled the pandemic.
+This is not just applicable for Russian data, but if other countries' data was missing, it would alter a major part of the analysis done above.
 
@@
 
-Weaknesses and next steps should also be included.
+While working on the paper we came across a few weaknesses that can be built on to make the report even more sound.
+Firstly, we could have tried to take into account the selection bias by choosing the countries except Russia randomly.
+It is obviously impractical to include all 195 countries in all plots, hence choosing a completely random subset of countries could have reduced the selection bias.
+However, it is also essential to note that it is preferable to compare Russia with countries with the same infrastructure and stature in the world, as that makes the comparisons more sound.
+Comparing Russian data with random countries might have led to this paper being less informative and less analytical.
+Then, another weakness was that while making @fig-map it was observed that the raw data file did not actually contain the GHS indices for all the countries around the world, for example, specifically Greenland was excluded from the dataset, which was the reason why some of the countries in the figure are shaded in grey showing that data was not available for them.
+If all the data was present for the figure, the analysis could have been even more sound.
+
+As a further scope, an extension to this report could include taking into account the proportion of population vaccinated for a country in 2021.
+Taking this factor into account, we would be able to clearly point out one of the prominent reasons why Russia might have lacked in their attempt to tackle the pandemic.
+Alternatively, we can flip the question for this paper around and delve deeper into the countries with the lowest GHS indices which surprisingly did better than expected.
+This would help us analyze the paper in positive notion, taking into account all the right things they did that led to them tackling the outbreak better than many well-prepared countries.
+
+In conclusion, we have observed that over 2020 and 2021, Russia failed to properly control the spread of the pandemic, and also effectively get its population vaccinated, which consequently lead to a large number of positive cases and deaths around the country.
+For future epidemics and outbreaks, Russia needs to make strict regulations and procedures which will help them contain the virus and not let it spread in the country and cause enormous amount of deaths.
 

### 2024-02-13 paper/paper.qmd
Added SSRP link. Removed unnecessary files from repo. Updated README.md.

@@
   - Janel Gilani
-thanks: "Code and data are available at: https://github.com/Ary4m3n/covid-effect-russia.git; Replication on Social Science Reproduction platform available at: "
+thanks: "Code and data are available at: https://github.com/Ary4m3n/covid-effect-russia.git; Replication on Social Science Reproduction platform available at: https://www.socialsciencereproduction.org/reproductions/8938b287-6681-4a5e-8993-42011a51be58/index"
 date: today

### 2024-02-13 paper/paper.qmd
Temp Push
Temp Push

--- 2024-02-12 paper/paper.qmd
@@
 date-format: long
-abstract: "Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Aenean tempus libero sit amet nisl dapibus, varius dictum magna pellentesque.
-In lacinia nec ligula eget condimentum.
-Aenean imperdiet enim sit amet felis volutpat, eu ornare ante pretium.
-Donec ullamcorper tempus facilisis.
-Mauris porttitor libero ultrices tortor facilisis, quis varius sapien ultricies.
-Pellentesque et metus vel quam mollis dignissim ultricies a dui.
-Cras sodales porttitor pulvinar.
-Proin vel nunc sit amet sapien venenatis pellentesque nec at felis."
+abstract: "GHS Index"
 format: pdf
@@
 #| message: false
-
+# install.packages("rnaturalearthdata")
 library(tidyverse)
@@
 Nuzzo and Jorge R.
-Ledesma's paper [@citeRepPaper] and findings and apply a Russian-facing lens to discuss the impact of the COVID-19 outbreak on Russia and what could have been done better and can be improved for the future in order for the impact of any upcoming pandemics can be lessened.
+Ledesma's paper [@citeRepPaper], where they used The Global Health Security (GHS) Index an indicator of how well prepared or "secure" a country is against epidemics or pandemics like the COVID-19 outbreak.
+We in-turn will use these findings and apply a Russian-facing lens to discuss the impact of the COVID-19 outbreak on Russia and what could have been done better and can be improved for the future in order for the impact of any upcoming pandemics can be lessened.
 Our paper aims to replicate their claim of how well-prepared countries did so poorly during the COVID-19 outbreak, where the Nuzzo and Ledesma have concentrated solely on the United States but in this paper we look at Russia with the same regards and analyse the impact of the pandemic on the death toll, and finally provide for ways to improve the laws, regulations and plans to tackle such outbreaks in the future.
@@
 The paper ends with the Discussion (@sec-discussion) section, where the findings of the paper have been analysed and delved deeper into by studying the measures that can be taken to prevent the country from experiencing such an extensive effect of future pandemics.
-Additionally, the limitations and a further scope for the paper has been discussed here.
+Additionally, the limitations and a further scope for the paper has been discussed here and concluded in the end.
 
@@
 R libraries and packages such as `tidyverse` [@citeTidyverse], `janitor` [@citeJanitor], `ggplot2` [@citeGgplot], `knitr` [@citeKnitr], `readr` [@citeReadr], `dplyr` [@citeDplyr], `rnaturalearth` [@citeRnaturalearth] and `sf` [@citeSf].
+Since, we have replicated the process of data cleaning and analysis, the data used in this paper is the same as the data used in the original paper.
+
 In the following sections, we will discuss the raw data in (@sec-data-raw), the data cleaning process in (@sec-data-clean) and then move on to discussing the measurement in (@sec-data-measurement).
@@
 The first raw dataset includes data on the Global Health Security index for all countries.
-The GHS index is an indicator of how well prepared or "secure" a country is against epidemics or pandemics like the COVID-19 outbreak.
 This dataset has a plethora of data on indices that are not relevant to us for this paper.
@@
 The second raw dataset includes data on all cause death rates for all countries from 1960-2021.
-This dataset has loads of information that was mainly unnecessary to us for this paper, mainly the data from 1960-2010 which is not required as we are only interested in looking at the effect of COVID-19 which is mainly data from 2020-2021.
-However, in order to study the trend, we will use data for all cause death rates for countries around the world from 2010-2021.
+Within this dataset, a plethora of information exists, the majority of which proved extraneous for the purposes of our paper.
+Specifically, data preceding the year 2010 is deemed unnecessary for our analysis, since we are only focusing on COVID-19 we will use this dataset to study the trend in death rates for countries around the world from 2010-2021.
 
-The third raw dataset is from the Institute of Health Metrics and Evaluation (IHME) which presents data on the COVID death rate, excess deaths, excess death rate and ratio of excess death rate over COVID death rate for all countries in the world.
-Again, this dataset too contained unnecessary information that was of no need to us which the cleaned data process outlined in @sec-data-clean.
+The third raw dataset originates from the Institute of Health Metrics and Evaluation (IHME) and provides information on COVID death rates, excess deaths, excess death rates, and the ratio of excess death rates to COVID death rates for all countries worldwide.
+Similar to the previous datasets, this dataset also contains extraneous information that is irrelevant to our study.
+However, we addressed this issue through the data cleaning process outlined in Section @sec-data-clean.
 
@@
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Aenean tempus libero sit amet nisl dapibus, varius dictum magna pellentesque.
-In lacinia nec ligula eget condimentum.
-Aenean imperdiet enim sit amet felis volutpat, eu ornare ante pretium.
-Donec ullamcorper tempus facilisis.
-Mauris porttitor libero ultrices tortor facilisis, quis varius sapien ultricies.
-Pellentesque et metus vel quam mollis dignissim ultricies a dui.
-Cras sodales porttitor pulvinar.
-Proin vel nunc sit amet sapien venenatis pellentesque nec at felis.
+When discussing COVID-19, (1) issues with underreported or missing data, particularly related to non-responses, are prevalent, (2) the lack of a standardized definition of COVID-19 deaths, which can lead to measurement errors (3) vastness of countries poses logistical challenges in reporting and conducting surveys, making it difficult to gather comprehensive data (4) variations in testing methodologies and availability of resources across regions can further distort the data.
+
+Many countries, especially developing nations like India, Brazil, and Russia, poses above issue.
+This poses a significant challenge in accurately measuring the pandemic's impact on a country.
+For example, the excess death rate, a variable in our analysis, measures the number of deaths occurring in a specific period above the expected number.
+This metric helps estimate the pandemic's impact.
+However, due to data truncation, the actual death rate isn't always precise, leading to inaccuracies in calculating the excess death rate.
+This introduces measurement errors, undermining the accuracy of both the actual death rate and the excess death rate.
+
+With respect to measurement and accounting for the GHS index, it encompasses various measurements grouped into six broad categories: (1) prevention, (2) detection and reporting, (3) rapid response, (4) health system, (5) compliance with international norms, and (6) risk environment.
+However, various measurement errors and qualitative biases can arise from the subjective nature of these categories.
+For example, the "prevention" category includes measurements such as "vaccination coverage" and "biosecurity and biosafety." However, the definition of "biosecurity and biosafety" can vary across countries, leading to measurement errors.
+Additionally, the GHS index is based on self-reported data, which can be subject to biases and inaccuracies.
+This can lead to measurement errors in the GHS index, which can further distort the analysis.
 

### 2024-02-13 paper/paper.qmd

--- 2024-02-13 paper/paper.qmd
@@
   - Janel Gilani
-thanks: "Code and data are available at: https://github.com/Ary4m3n/covid-effect-russia.git; Replication on Social Science Reproduction platform available at: "
+thanks: "Code and data are available at: https://github.com/Ary4m3n/covid-effect-russia.git; Replication on Social Science Reproduction platform available at: https://www.socialsciencereproduction.org/reproductions/8938b287-6681-4a5e-8993-42011a51be58/index"
 date: today
 date-format: long
-abstract: "Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Aenean tempus libero sit amet nisl dapibus, varius dictum magna pellentesque.
-In lacinia nec ligula eget condimentum.
-Aenean imperdiet enim sit amet felis volutpat, eu ornare ante pretium.
-Donec ullamcorper tempus facilisis.
-Mauris porttitor libero ultrices tortor facilisis, quis varius sapien ultricies.
-Pellentesque et metus vel quam mollis dignissim ultricies a dui.
-Cras sodales porttitor pulvinar.
-Proin vel nunc sit amet sapien venenatis pellentesque nec at felis."
+abstract: "GHS Index"
 format: pdf

### 2024-02-13 paper/paper.qmd
Abstract and Measurement
- Added Abstract
- Added measurement sections.

@@
 date-format: long
-abstract: "GHS Index"
+abstract: " Despite Russia being deemed well-prepared for pandemics, it faced an unprecedented rise in death rates compared to other nations.
+This study aims to replicate Nuzzo and Ledesma's findings, examining COVID-19's impact on Russia while considering variables such as the Global Health Security (GHS) Index and excess death rates.
+Our analysis revealed a significantly elevated excess death rate in Russia, contrasting with the lower rates in more prepared countries.
+This research sheds light on the COVID-19 outbreak's impact on Russia and identifies areas for improvement for well-prepared countries to mitigate future pandemics' effects, emphasizing the importance of better planning and regulations."
 format: pdf
@@
 
-When discussing COVID-19, (1) issues with underreported or missing data, particularly related to non-responses, are prevalent, (2) the lack of a standardized definition of COVID-19 deaths, which can lead to measurement errors (3) vastness of countries poses logistical challenges in reporting and conducting surveys, making it difficult to gather comprehensive data (4) variations in testing methodologies and availability of resources across regions can further distort the data.
+With respect to measurement and accounting for the GHS index, it encompasses various measurements grouped into six broad categories: (1) prevention, (2) detection and reporting, (3) rapid response, (4) health system, (5) compliance with international norms, and (6) risk environment.
+These values are calculated based on the data available for each country which are self-reported and are then used to calculate the overall GHS index [@GHSIndexMethodology].`
 
-Many countries, especially developing nations like India, Brazil, and Russia, poses above issue.
+In addition to tracking deaths, a crucial data point during the COVID-19 pandemic is the excess death rate, representing the number of deaths occurring within a specific period above the expected baseline.
+This metric serves as a key indicator of the pandemic's impact on mortality.
+In our dataset, death rates are standardized to a per 100,000 population basis, facilitating meaningful comparisons across countries.
+
+The process of collecting this data is primarily overseen by institutions like the Institute for Health Metrics and Evaluation (IHME), which collates information from various sources available for each country.
+Deaths included in the excess death rate calculation encompass those directly attributed to COVID-19, as well as those indirectly influenced by the pandemic where COVID-19 appeared on the death certificate.
+For instance, deaths resulting from delayed treatment due to healthcare system strain caused by the pandemic are also accounted for.
+
+Observation of deaths involves various mechanisms.
+Initially, healthcare facilities and authorities report COVID-19-related deaths through diagnostic tests and medical records.
+Deaths with an underlying cause of death of COVID-19 are not included in these estimates.
+For example, in some cases, COVID-19 may have contributed to the death, but the underlying cause of death was another cause, such as terminal cancer.
+Additionally, mortality statistics are compiled from death certificates and official registries.
+Tallying these deaths are then put into the Government's official records.
+
+Once observed deaths are verified and categorized, they are tallied according to established criteria for inclusion as defined by IHME.
+the excess death rate calculation.
+This meticulous approach ensures the accuracy and reliability of the data used to compute the excess death rate for each country.
+
+However, due to data truncation, the actual death rate isn't always precise, leading to inaccuracies in calculating the excess death rate.
+This introduces measurement errors, undermining the accuracy of both the actual death rate and the excess death rate.
+This is a significant challenge in accurately measuring the pandemic's impact on a country.
+
+<!-- When discussing COVID-19, (1) issues with underreported or missing data, particularly related to non-responses, are prevalent, (2) the lack of a standardized definition of COVID-19 deaths, which can lead to measurement errors (3) vastness of countries poses logistical challenges in reporting and conducting surveys, making it difficult to gather comprehensive data (4) variations in testing methodologies and availability of resources across regions can further distort the data.
+-->
+
+<!-- Many countries, especially developing nations like India, Brazil, and Russia, poses above issue.
 This poses a significant challenge in accurately measuring the pandemic's impact on a country.
@@
 This introduces measurement errors, undermining the accuracy of both the actual death rate and the excess death rate.
+-->
 
-With respect to measurement and accounting for the GHS index, it encompasses various measurements grouped into six broad categories: (1) prevention, (2) detection and reporting, (3) rapid response, (4) health system, (5) compliance with international norms, and (6) risk environment.
+<!-- With respect to measurement and accounting for the GHS index, it encompasses various measurements grouped into six broad categories: (1) prevention, (2) detection and reporting, (3) rapid response, (4) health system, (5) compliance with international norms, and (6) risk environment.
 However, various measurement errors and qualitative biases can arise from the subjective nature of these categories.
@@
 This can lead to measurement errors in the GHS index, which can further distort the analysis.
+-->
 
@@
 
+It should be noted that different countries employ varying methods of reporting their data, including mortality rates and excess death rates.
+Therefore, this study primarily relied on observational data, but it's crucial to account for differential biases in the data.
+For instance, some countries might underreport their deaths, while others might overreport them.
+Such discrepancies directly impact the analysis conducted in this paper.
+
+Furthermore, COVID-19 policies were predominantly implemented at the state and local levels, introducing variability throughout the pandemic.
+Differences in policy implementation among these communities, coupled with the presence of unmeasured differences, complicate the ability to draw causal inferences.
+
 There are a few ethical implications that we found while working on the paper.

### 2024-02-13 paper/paper.qmd
Simulation and Tests added. Edited 02-data_cleaning.R and 99-replication.R. Minor edits to README.md. paper.qmd recompiled.

@@
 With respect to measurement and accounting for the GHS index, it encompasses various measurements grouped into six broad categories: (1) prevention, (2) detection and reporting, (3) rapid response, (4) health system, (5) compliance with international norms, and (6) risk environment.
-These values are calculated based on the data available for each country which are self-reported and are then used to calculate the overall GHS index [@GHSIndexMethodology].`
+These values are calculated based on the data available for each country which are self-reported and are then used to calculate the overall GHS index [@GHSIndexMethodology].
 

### 2024-02-13 scripts/00-simulate_data.R

@@ -1,19 +1,52 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Simulates the data tables shown in the other/sketches/datasets.pdf. There are 3 tables simulated here as mentioned below.
+# Author: Aryaman Sharma
+# Date: 13 February 2024
+# Contact: aryaman.sharma@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: None
+# Any other information needed?: Run this script to get an idea of how the data tables will look like.
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(janitor)
+library(countries) # read documentation: https://cran.r-project.org/web/packages/countries/readme/README.html#:~:text=countries%20is%20an%20R%20package,and%20easily%20make%20world%20maps.
 
 #### Simulate data ####
-# [...ADD CODE HERE...]
+# We set the seed here to help make this script reproducible
+set.seed(43)
 
+# This piece of code generates a table that contains GHS indicies for each of the 195 countries in the world
+simulated_country_ghs_data <- 
+  tibble(
+    "Country Name" = random_countries(195),
+    "GHS Index" = runif(n = 195, min = 16, max = 76)
+  )
 
+# This piece of code generates a table that contains all cause death rates for 4 countries between 2010 and 2021
+simulated_all_cause_death_rate_data <-
+  tibble(
+    "Country Name" = random_countries(4),
+    "2010" = runif(n = 4, min = 4, max = 20),
+    "2011" = runif(n = 4, min = 4, max = 20),
+    "2012" = runif(n = 4, min = 4, max = 20),
+    "2013" = runif(n = 4, min = 4, max = 20),
+    "2014" = runif(n = 4, min = 4, max = 20),
+    "2015" = runif(n = 4, min = 4, max = 20),
+    "2016" = runif(n = 4, min = 4, max = 20),
+    "2017" = runif(n = 4, min = 4, max = 20),
+    "2018" = runif(n = 4, min = 4, max = 20),
+    "2019" = runif(n = 4, min = 4, max = 20),
[…]

### 2024-02-13 scripts/02-data_cleaning.R

@@ -1,11 +1,11 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Cleans the raw data obtained from Nuzzo and Ledesma's paper and saves it to the directory "data/analysis_data"
+# Author: Aryaman Sharma
+# Date: 10 February 2024
+# Contact: aryaman.sharma@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: None
+# Any other information needed?
 
 #### Workspace setup ####
 library(ggplot2)

### 2024-02-13 scripts/03-test_data.R

@@ -1,15 +1,105 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Tests for the cleaned dataset to check if the requirements match that of the simulated data.
+# Author: Aryaman Sharma
+# Date: 13 February 2024
+# Contact: aryaman.sharma@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Kindly run 00-simulate_data.R and 02-data_cleaning.R before running this file.
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
 
 #### Test data ####
+# Let us first work on the GHS Index data for each country
+test_ghs_index_data_2021 <- read_csv("data/analysis_data/cleaned_GHS_index.csv", show_col_types = FALSE)
+# First Test is to check if there are 195 countries in total here
+test_ghs_index_data_2021$country |>
+  length() == 195
+# Second Test is the check if all the GHS index values are non-negative
+test_ghs_index_data_2021$overall_score |>
+  min() >= 0
+# Third Test is to check if the country column is of class "character"
+test_ghs_index_data_2021$country |>
+  class() == "character"
+# Fourth Test is to check if the ghs index column is of class "numeric"
+test_ghs_index_data_2021$overall_score |>
+  class() == "numeric"
+
+# Let us now work on the all cause death data for 4 countries for 2010-2021
+test_deaths_data <- read_csv("data/analysis_data/cleaned_all_cause_dth_rates.csv", show_col_types = FALSE)
+# First Test is to check if there are 4 countries being analysed
+test_deaths_data$country_name |>
+  length() == 4
+# Second Test is to check if the country column is of class "character"
+test_deaths_data$country_name |>
+  class() == "character"
+# Third Test is a set of tests to check if the all cause death rate for 2010-2021 is non-negative
+test_deaths_data$`2010` |>
+  min() >= 0
+test_deaths_data$`2011` |>
+  min() >= 0
+test_deaths_data$`2012` |>
+  min() >= 0
+test_deaths_data$`2013` |>
[…]

### 2024-02-13 scripts/99-replications.R

@@ -1,11 +1,11 @@
 #### Preamble ####
-# Purpose: Replicated graphs from... [...UPDATE THIS...]
+# Purpose: Replicated graphs from Nuzzo and Ledesma's paper provided in "other/literature"
 # Author: Aryaman Sharma
 # Date: 10 February 2024
 # Contact: aryaman.sharma@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Run 02-data_cleaning.R before running this script.
+# Any other information needed?: This script will produce 5 graphs in total, which is Figure 1, Figure 2a, Figure 2b, Figure 3a and Figure 3b of Nuzzo and Ledesma's paper
 
 
 #### Workspace setup ####

### 2024-02-14 paper/paper.qmd
New Sections and Feedback Draft
- Added new section based on peer review

@@
 date-format: long
-abstract: " Despite Russia being deemed well-prepared for pandemics, it faced an unprecedented rise in death rates compared to other nations.
+abstract: "Despite Russia being deemed well-prepared for pandemics, it faced an unprecedented rise in death rates compared to other nations.
 This study aims to replicate Nuzzo and Ledesma's findings, examining COVID-19's impact on Russia while considering variables such as the Global Health Security (GHS) Index and excess death rates.
@@
 Since, we have replicated the process of data cleaning and analysis, the data used in this paper is the same as the data used in the original paper.
+The estimands for this analysis are preparedness of countries to tackle pandemic, specifically Russia, and their degree of success in controlling the pandemic.
+To estimate these, we utilize metrics such as the Global Health Security (GHS) Index and the excess death rate respectively.
 
@@
 
-In addition to tracking deaths, a crucial data point during the COVID-19 pandemic is the excess death rate, representing the number of deaths occurring within a specific period above the expected baseline.
-This metric serves as a key indicator of the pandemic's impact on mortality.
-In our dataset, death rates are standardized to a per 100,000 population basis, facilitating meaningful comparisons across countries.
-
-The process of collecting this data is primarily overseen by institutions like the Institute for Health Metrics and Evaluation (IHME), which collates information from various sources available for each country.
-Deaths included in the excess death rate calculation encompass those directly attributed to COVID-19, as well as those indirectly influenced by the pandemic where COVID-19 appeared on the death certificate.
+A crucial data point for the COVID-19 pandemic is the excess death rate, representing the number of deaths occurring within a specific period above the expected baseline.
+The process of collecting this data is primarily overseen by Institute for Health Metrics and Evaluation (IHME), which collates information from various sources available for each country.
+Deaths included in the excess death rate calculation consists of those directly attributed to COVID-19, or in which COVID-19 appeared on the death certificates.
 For instance, deaths resulting from delayed treatment due to healthcare system strain caused by the pandemic are also accounted for.
@@
 
-Once observed deaths are verified and categorized, they are tallied according to established criteria for inclusion as defined by IHME.
-the excess death rate calculation.
-This meticulous approach ensures the accuracy and reliability of the data used to compute the excess death rate for each country.
+However, discussions surrounding COVID-19 present several challenges, including: (1) issues related to underreported or missing data, which can occur due to societal stigma at the local level, or countries intentionally concealing data to portray preparedness; (2) the absence of a universally accepted definition for COVID-19 deaths, leading to potential measurement inaccuracies; (3) the vast geographical expanse of countries, posing logistical hurdles in data collection and reporting efforts; and (4) disparities in testing methodologies and resource availability across regions, further complicating data interpretation.
 
-However, due to data truncation, the actual death rate isn't always precise, leading to inaccuracies in calculating the excess death rate.
-This introduces measurement errors, undermining the accuracy of both the actual death rate and the excess death rate.
-This is a significant challenge in accurately measuring the pandemic's impact on a country.
-
-<!-- When discussing COVID-19, (1) issues with underreported or missing data, particularly related to non-responses, are prevalent, (2) the lack of a standardized definition of COVID-19 deaths, which can lead to measurement errors (3) vastness of countries poses logistical challenges in reporting and conducting surveys, making it difficult to gather comprehensive data (4) variations in testing methodologies and availability of resources across regions can further distort the data.
--->
-
-<!-- Many countries, especially developing nations like India, Brazil, and Russia, poses above issue.
-This poses a significant challenge in accurately measuring the pandemic's impact on a country.
-For example, the excess death rate, a variable in our analysis, measures the number of deaths occurring in a specific period above the expected number.
-This metric helps estimate the pandemic's impact.
-However, due to data truncation, the actual death rate isn't always precise, leading to inaccuracies in calculating the excess death rate.
-This introduces measurement errors, undermining the accuracy of both the actual death rate and the excess death rate.
--->
-
-<!-- With respect to measurement and accounting for the GHS index, it encompasses various measurements grouped into six broad categories: (1) prevention, (2) detection and reporting, (3) rapid response, (4) health system, (5) compliance with international norms, and (6) risk environment.
-However, various measurement errors and qualitative biases can arise from the subjective nature of these categories.
-For example, the "prevention" category includes measurements such as "vaccination coverage" and "biosecurity and biosafety." However, the definition of "biosecurity and biosafety" can vary across countries, leading to measurement errors.
-Additionally, the GHS index is based on self-reported data, which can be subject to biases and inaccuracies.
-This can lead to measurement errors in the GHS index, which can further distort the analysis.
--->
+Many countries, particularly nations such as India, Brazil, and Russia, which had the highest death count, faced the aforementioned challenges.
+This presents a significant obstacle in accurately gauging the pandemic's toll within a country.
+For instance, the excess death rate, a key metric in our analysis, serves to estimate the pandemic's impact.
+However, due to data truncation and measurement errors, the actual death rate may not always be precisely captured.
+Consequently, inaccuracies arise in calculating the excess death rate, introducing measurement errors that undermine the reliability of both the actual death rate and the excess death rate in data analysis.
 
@@
 
-This section will be divided into three parts where all together we aim to look at Russia's capabilty according to the GHS index to tackle pandemics (@sec-results-ghs), then compare Russia's trend in all cause death rates and excess death rates with those of other countries like India, South Korea and the United states (@sec-results-death-rates) and finally study the relationship between the excess death rate and GHS index for Russia along with a subset of countries in the world (@sec-results-death-rates).
+The results section is divided into four parts where all together we aim to look at Russia's capability according to the GHS index to tackle pandemics (@sec-results-ghs), then compare Russia's death rates trends with those of other countries like India, South Korea and the United states (@sec-results-death-rates), then using those trends we dive into Russia's excess death rates while comparing them with other countries (@sec-results-excess-rates) and finally study the relationship between the excess death rate and GHS index for Russia along with a subset of countries in the world (@sec-results-relationship).
 
@@
 
+## Excess Death Rates {#sec-results-excess-rates}
 This observation leads us to @fig-excess-death-rate, where we present the excess death rate for the four countries for 2020 and 2021.
@@
 
-## Relationship between the Excess Death Rate and Global Health Security Index (GHS) {#sec-results-death-rates}
+## Relationship between the Excess Death Rate and Global Health Security Index (GHS) {#sec-results-relationship}
 

### 2024-02-14 paper/paper.qmd
proof-reading changes (until discussions)

@@
 ---
-title: "Why Did Russia Fare So Poorly during COVID: An Analysis of Why One of the Best Prepared Countries Could Not Control Excess Deaths"
+title: "Evaluating Russia's Preparedness and Response to the COVID-19 Pandemic: An Analysis of Excess Death Rates and the Global Health Security Index"
 author: 
@@
 date-format: long
-abstract: "Despite Russia being deemed well-prepared for pandemics, it faced an unprecedented rise in death rates compared to other nations.
-This study aims to replicate Nuzzo and Ledesma's findings, examining COVID-19's impact on Russia while considering variables such as the Global Health Security (GHS) Index and excess death rates.
-Our analysis revealed a significantly elevated excess death rate in Russia, contrasting with the lower rates in more prepared countries.
-This research sheds light on the COVID-19 outbreak's impact on Russia and identifies areas for improvement for well-prepared countries to mitigate future pandemics' effects, emphasizing the importance of better planning and regulations."
-format: pdf
+abstract: "This study investigates the unexpected surge in death rates during the COVID-19 pandemic in Russia, a country considered well-prepared according to the Global Health Security (GHS) Index.
+Building upon Nuzzo and Ledesma's work, we analyze the impact of COVID-19 in Russia, incorporating factors such as the GHS Index and excess death rates.
+Our findings reveal a substantial increase in excess death rates in Russia, diverging from the lower rates observed in nations deemed well-prepared.
+This research contributes to our understanding of the repercussions of the COVID-19 outbreak in Russia, highlighting areas for improvement in prepared countries to mitigate the effects of future pandemics."
+format: 
+  pdf:
+    tbl-pos: "h"
+    fig-pos: "h"
 toc: true
@@
 
-The COVID-19 pandemic has led to an enormous loss of life over the world and has presented an unprecedented challenge to public health, foreign policy and has additionally had an adverse effect on countries' economies around the world, over the past 5 years [@citeWHOeffect].
-There have been over 774 million reported confirmed cases of the virus being contract and the outbreak has caused over 7 million deaths, which are claimed to be "under counted" and "under reported", suggesting that the actual number is even higher around 18.2 - 33.5 million based on the excess mortality estimates [@citeWikiCOVID].
-This excessively high number of deaths has alerted the world to assess how countries assumed to be well-prepared to tackle such outbreaks were affected to this extent.
+The global impact of the COVID-19 pandemic on public health, foreign policy, and economies has been profound, resulting in an unprecedented loss of life worldwide over the past five years (World Health Organization [@citeWHOeffect]).
+With over 774 million confirmed cases and more than 7 million reported deaths, these figures, considered to be undercounted, suggest an actual death toll ranging from 18.2 to 33.5 million based on excess mortality estimates (WikiCOVID [@citeWikiCOVID]).
+The staggering number of deaths has prompted a critical examination of countries presumed to be well-prepared to handle such outbreaks and the factors contributing to their extensive impact.
 
-There were a subset of countries, namely India, the United States of America, Brazil, Russia and Mexico which were impacted the most in terms of having the highest cumulative total deaths [@citeThinkGlobalHealth].
-According to the global data, Russia's COVID-19 related deaths since the start of the pandemic have been over 230,000.
-Russia also holds the highest mortality rate in Europe and the second-highest rate in Asia [@citeCarnegieRussia].
-For these reasons and more, Russia and its residents have suffered a lot over the past few years, wherein the adverse impact of the outbreak has lead to a plethora of impacts on the country's economy and standard of living.
+Among the most severely affected nations, including India, the United States of America, Brazil and Mexico, Russia, a nation deemed well-prepared according to various metrics, was specifically chosen for this study due to its unexpected and disproportionately high death toll in the face of the pandemic.
+With over 230,000 reported COVID-19-related deaths, Russia holds the highest mortality rate in Europe and the second-highest in Asia (ThinkGlobalHealth [@citeThinkGlobalHealth]; Carnegie Russia [@citeCarnegieRussia]).
+This selection aims to uncover unique insights into the challenges faced by ostensibly prepared nations.
 
-This paper will follow a reproduction of Jennifer B.
+This paper replicates the methodology employed by Jennifer B.
 Nuzzo and Jorge R.
-Ledesma's paper [@citeRepPaper], where they used The Global Health Security (GHS) Index an indicator of how well prepared or "secure" a country is against epidemics or pandemics like the COVID-19 outbreak.
-We in-turn will use these findings and apply a Russian-facing lens to discuss the impact of the COVID-19 outbreak on Russia and what could have been done better and can be improved for the future in order for the impact of any upcoming pandemics can be lessened.
-Our paper aims to replicate their claim of how well-prepared countries did so poorly during the COVID-19 outbreak, where the Nuzzo and Ledesma have concentrated solely on the United States but in this paper we look at Russia with the same regards and analyse the impact of the pandemic on the death toll, and finally provide for ways to improve the laws, regulations and plans to tackle such outbreaks in the future.
-Our reproduction found out that how ever well prepared Russia seemed to be, it experienced an unprecedented rise in the death rate compared to other countries, and its excess death rate was one of the highest amongst all countries.
-This leads us to make an important judgement that better planning and regulations can help reduce the impact of such outbreaks in the future, which will be studied in this paper.
+Ledesma in their study, utilizing the Global Health Security (GHS) Index as an indicator of a country's preparedness against epidemics or pandemics like COVID-19 (RepPaper [@citeRepPaper]).
+The choice of GHS is deliberate, considering its comprehensive evaluation of various health security capacities.
+This metric offers a nuanced understanding of a country's ability to prevent, detect, and respond to public health emergencies, making it a suitable choice over other metrics.
+However, our focus extends to Russia, providing insights into the impact of the outbreak and proposing improvements for future pandemic mitigation efforts.
+Despite Russia's apparent preparedness, our reproduction reveals an unprecedented rise in the death rate compared to other countries, with one of the highest excess death rates globally.
+This prompts a crucial assessment that better planning and regulations are imperative to mitigate the impact of such outbreaks, a key aspect explored in this paper.
 
-This paper is structured using the following sections: Data, Results and Discussion.
-In the Data (@sec-data) section, the data source of the datasets from the paper being replicated [@citeRepPaper] is discussed and the measurement and data cleaning process is outlined.
-In the Results (@sec-results) section, the paper summarizes the data findings and presents relevant maps and plots in order to study the trends.
-The paper ends with the Discussion (@sec-discussion) section, where the findings of the paper have been analysed and delved deeper into by studying the measures that can be taken to prevent the country from experiencing such an extensive effect of future pandemics.
-Additionally, the limitations and a further scope for the paper has been discussed here and concluded in the end.
+The structure of this paper comprises three sections: Data, Results, and Discussion.
+In the Data section (@sec-data), we discuss the data source and the process of measuring and cleaning the datasets from the replicated paper (RepPaper [@citeRepPaper]).
+The Results section (@sec-results) summarizes the findings, presenting maps and plots to illustrate trends.
+The Discussion section (@sec-discussion) delves into the analysis of our findings, exploring measures to prevent extensive effects of future pandemics on the country.
+Additionally, we discuss limitations and propose avenues for further research, concluding the paper.
 
@@
 
-As mentioned above, the data analysed in this paper is from the reproduction package of the paper by Nuzzo and Ledesma [@citeRepPaper].
-The data was cleaned and analysed using the open source `R` programming language [@citeR].
-R libraries and packages such as `tidyverse` [@citeTidyverse], `janitor` [@citeJanitor], `ggplot2` [@citeGgplot], `knitr` [@citeKnitr], `readr` [@citeReadr], `dplyr` [@citeDplyr], `rnaturalearth` [@citeRnaturalearth] and `sf` [@citeSf].
-Since, we have replicated the process of data cleaning and analysis, the data used in this paper is the same as the data used in the original paper.
-The estimands for this analysis are preparedness of countries to tackle pandemic, specifically Russia, and their degree of success in controlling the pandemic.
-To estimate these, we utilize metrics such as the Global Health Security (GHS) Index and the excess death rate respectively.
+The dataset under examination in this study originates from the replication package associated with Nuzzo and Ledesma's paper [@citeRepPaper].
+Employing the open-source `R` programming language [@citeR], we conducted the cleaning and analysis procedures, leveraging several R libraries and packages such as `tidyverse` [@citeTidyverse], `janitor` [@citeJanitor], `ggplot2` [@citeGgplot], `knitr` [@citeKnitr], `readr` [@citeReadr], `dplyr` [@citeDplyr], `rnaturalearth` [@citeRnaturalearth], and `sf` [@citeSf].
 
-In the following sections, we will discuss the raw data in (@sec-data-raw), the data cleaning process in (@sec-data-clean) and then move on to discussing the measurement in (@sec-data-measurement).
+Our data cleaning and analysis processes replicate those of the original paper, ensuring consistency in the datasets used.
+In terms of estimand, the primary focus of this analysis is to assess the preparedness of countries, specifically Russia, in addressing pandemics, and to gauge their success in controlling the associated impact.
+Key estimators include the Global Health Security (GHS) Index, offering insights into overall preparedness, and the excess death rate, serving as an indicator of the pandemic's impact.
+
+The subsequent sections will provide a comprehensive overview of the raw data in (@sec-data-raw), delve into the intricacies of the data cleaning process in (@sec-data-clean), and subsequently, explore the measurement aspects in (@sec-data-measurement).
 
@@
 
-This paper will replicate the data that was obtained from the replication package of the paper [@citeRepPaper].
-Specifically, in the paper we use 3 datasets that will help us analyse the claim of Russia not being able to control the wave of the COVID-19 pandemic well.
-In general, there is a lot of data available on the effect of COVID-19 on countries, however, all the data we needed for this paper was included in the paper by Nuzzo and Ledesma.
+This study replicates the data sourced from the replication package of Nuzzo and Ledesma's paper [@citeRepPaper].
+Specifically, we utilize three datasets to investigate the assertion that Russia struggled to effectively manage the surge of the COVID-19 pandemic.
+Despite the wealth of data available on the global impact of COVID-19, the comprehensive datasets provided by Nuzzo and Ledesma serve as the primary source for our analysis.
 
-The first raw dataset includes data on the Global Health Security index for all countries.
-This dataset has a plethora of data on indices that are not relevant to us for this paper.
-Additionally, the dataset contains data for years 2019 and 2021.
-For our analysis, we only require data for the GHS index of countries for 2021.
+The first raw dataset encompasses the Global Health Security index for all countries, offering a wide array of indices, some of which are not pertinent to our study.
+Notably, this dataset spans the years 2019 and 2021; however, our analysis focuses exclusively on the GHS index for the year 2021.
 
-The second raw dataset includes data on all cause death rates for all countries from 1960-2021.
-Within this dataset, a plethora of information exists, the majority of which proved extraneous for the purposes of our paper.
-Specifically, data preceding the year 2010 is deemed unnecessary for our analysis, since we are only focusing on COVID-19 we will use this dataset to study the trend in death rates for countries around the world from 2010-2021.
+The second raw dataset provides data on all-cause death rates for all countries from 1960 to 2021.
+The "all-cause death rate" refers to the overall mortality rate in a given population, considering deaths from all possible causes.
+It is a comprehensive measure that includes deaths from various diseases, accidents, and other factors.
+While this dataset contains extensive information, we restrict our analysis to data from 2010 onwards, as earlier years are deemed unnecessary for our COVID-19-focused investigation.
 
-The third raw dataset originates from the Institute of Health Metrics and Evaluation (IHME) and provides information on COVID death rates, excess deaths, excess death rates, and the ratio of excess death rates to COVID death rates for all countries worldwide.
-Similar to the previous datasets, this dataset also contains extraneous information that is irrelevant to our study.
-However, we addressed this issue through the data cleaning process outlined in Section @sec-data-clean.
+The third raw dataset, sourced from the Institute of Health Metrics and Evaluation (IHME), provides us information on COVID death rates, excess deaths, excess death rates, and the ratio of excess death rates to COVID death rates for countries globally.
+Similar to the preceding datasets, unnecessary information was present, requiring thorough data cleaning, a process detailed in Section @sec-data-clean.
 
-In the next section (@sec-data-clean), we will outline the data-cleaning process and also show the structure of the cleaned data.
+The subsequent section (@sec-data-clean) will show the data cleaning process, elucidating the structure of the refined dataset employed in our analysis.
 
@@
 
-As stated above in @sec-data-raw, we start with the first dataset on GHS indices for all countries in the world.
-@tbl-cleaned-ghs shows the first 6 countries and their GHS indices.
-Here we only have 2 variables, namely the Country Name and the respective GHS index.
-We cleaned this dataset by selecting only the two columns for the respective variables and filtering the data for the year of 2021.
+As outlined in @sec-data-raw, our initial step involves the first dataset concerning Global Health Security (GHS) indices for all countries globally.
+Presented in @tbl-cleaned-ghs are the first six countries along with their corresponding GHS indices.
+The dataset has been refined to include only two variables, namely Country Name and the respective GHS index.
+Cleaning involved the selection of these two columns and filtering the data specifically for the year 2021.
 
@@
 #| tbl-cap: "Cleaned Data showing GHS indices for first 6 countries"
-#| tbl-cap-location: bottom
 #| warning: false
@@
 
-Second, we have a raw dataset on all cause death rates for all countries from 1960-2021.
-We cleaned this dataset by selecting a subset of countries that helped us analyse how Russia fared against other countries in death ratesover the years.
-@tbl-cleaned-death-rates shows the structure of the cleaned data after we filtered it for the time period of 2010-2021 for India, South Korea, Russia and the United States.
+Moving to the second dataset covering all-cause death rates for countries from 1960 to 2021, the cleaned version is showcased in @tbl-cleaned-death-rates.
+To analyze Russia's position relative to other nations in terms of death rates, we filtered this dataset for the years 2010-2021, focusing on countries like India, South Korea, Russia, and the United States.
+The refined dataset structure is highlighted, enabling a clearer examination of death rate trends.
 
@@
 #| tbl-cap: "Cleaned Data showing all cause death rates for 4 countries"
-#| tbl-cap-location: bottom
 #| warning: false
@@
 
-The final raw dataset, as mentioned in @sec-data-raw, that we have is the IHME data on the COVID death rate, excess deaths, excess death rate and ratio of excess death rate over COVID death rate for all countries in the world.
-In this paper we only needed the two variables excess death rate and the country name, however for better analysis of Russia's ability to control the pandemic's effect, we combined this data with the respective GHS index as shown in @tbl-cleaned-combined.
-Hence, there are three variables, the country name, and the respective excess death rate and GHS index.
+The last raw dataset, detailed in @sec-data-raw, originates from the Institute of Health Metrics and Evaluation (IHME), providing data on COVID death rates, excess deaths, excess death rates, and the ratio of excess death rates to COVID death rates for countries worldwide.
+While our primary interest was in the excess death rate and the country name, for a more comprehensive analysis of Russia's pandemic management, we integrated this data with the respective GHS index, showcased in @tbl-cleaned-combined.
+Consequently, this dataset comprises three variables: the country name, the respective excess death rate, and the GHS index.
 
@@
 #| tbl-cap: "Cleaned Data showing Excess Death Rate and GHS index for first 6 chosen countries"
-#| tbl-cap-location: bottom
 #| warning: false
@@
 
-Now that we have cleaned our raw datasets and have generated clean tables for analysis, we will go on to the Results Section (@sec-results) and delve deeper into studying the effect of the COVID-19 pandemic on Russia and if it was impacted significantly more than other countries.
+With the completion of the data cleaning phase and the generation of refined tables for analysis, our next section will delve into the Measurement (@sec-data-measurement), offering an in-depth exploration of how the phenomena observed translated into entries in our dataset.
+Understanding this process is crucial for interpreting patterns and drawing meaningful conclusions from the data analysis results.
 
@@
 
+The subsequent section (@sec-results) will present the results of our analysis, offering insights into Russia's preparedness and response to the COVID-19 pandemic, and the implications of the findings.
+
 # Results {#sec-results}
 
-The results section is divided into four parts where all together we aim to look at Russia's capability according to the GHS index to tackle pandemics (@sec-results-ghs), then compare Russia's death rates trends with those of other countries like India, South Korea and the United states (@sec-results-death-rates), then using those trends we dive into Russia's excess death rates while comparing them with other countries (@sec-results-excess-rates) and finally study the relationship between the excess death rate and GHS index for Russia along with a subset of countries in the world (@sec-results-relationship).
+The results section is divided into four parts where all together we aim to look at Russia's capability according to the GHS index to tackle pandemics (@sec-results-ghs), then compare Russia's all-causes mortality rate trends with those of other countries like India, South Korea and the United states (@sec-results-death-rates), then using those trends we dive into Russia's excess death rates while comparing them with other countries (@sec-results-excess-rates) and finally study the relationship between the excess death rate and GHS index for Russia along with a subset of countries in the world (@sec-results-relationship).
 
@@
 
-As explained in @sec-data-raw, the GHS index is an indicator of how well prepared a country is to tackle epidemics or pandemics like the COVID-19 outbreak.
-The GHS index gives one an idea of how the sound health care system  as well as the rules and regulations of a certain country are in the context of facing pandemics.
-Generally, the higher the GHS index of a country, the more well-prepared and secure they are.
-@fig-map shows a representation of the world map with countries shaded on the basis of their GHS indices.
-This map is a replication of figure 1 in Nuzzo and Ledesma paper [@citeRepPaper] that is being replicated in this report.
+As outlined in @sec-data-raw, the Global Health Security (GHS) index serves as a metric indicating a country's preparedness to address epidemics or pandemics, such as the COVID-19 outbreak.
+Reflecting both the robustness of the healthcare system and the regulatory framework, a higher GHS index suggests a more secure and well-prepared nation.
+@fig-map presents a world map illustrating countries shaded based on their GHS indices.
+This replication of figure 1 from Nuzzo and Ledesma's paper [@citeRepPaper] provides an overview of global preparedness.
 
@@
 
-We can observe here that there are countries like the United States and Australia whose GHS indices are one of the highest, implying that they were deemed extremely well prepared for epidemics.
-Then we see countries like Russia and China who were, although moderately when compared to the aforementioned countries, but still well prepared to tackle pandemics.
-Finally, we see countries such as Venezuela and Yemen which were said to be the least secure of all.
+Noteworthy observations reveal countries like the United States and Australia boasting some of the highest GHS indices, indicating exceptional preparedness for epidemics.
+Countries like Russia and China, while moderately prepared compared to the former, still exhibit a substantial level of readiness.
+Conversely, nations such as Venezuela and Yemen rank among the least secure.
 
-@fig-map provides this report with a background on how secure Russia was in tackling epidemics and pandemics, which as seen in the map, it was in a good position to face such outbreaks where numerous countries were deemed less prepared than Russia.
-This will aid us in drawing meaningful analyses throughout the paper, specifically in @sec-results-death-rates, sec-results-death-rates and @sec-discussion.
+@fig-map provides us essential contextual information on Russia's preparedness in the face of epidemics and pandemics.
+Positioned favorably, Russia emerges as well-prepared compared to numerous countries deemed less secure.
+This background sets the stage for meaningful analyses throughout the paper, particularly in @sec-results-death-rates and @sec-discussion.
 
@@
 
-Now we move on to analyzing the trends in the all cause death rates per 1000 population over 2010-2021, which includes COVID deaths as well as general deaths which were not caused by the outbreak, and the trend in the excess death rates per 100 population in 2020 and 2021.
-The excess death rates in this section were calculated by finding a difference between the expected death rate and the actual death rate for that year.
-
-@fig-all-cause-death-rate shows us the trend for the all cause death rates per 1000 population over 2010-2021 for India, South Korea, Russia and the United States using the data presented in @tbl-cleaned-death-rates.
-We observe here that from the years 2010-2019, there is a general gradually decreasing trend in the death rates in the case of Russia and India, and a gradually increasing trend in the death rates in the case of South Korea and the United States.
-The dashed line represents the line of best fit for the trend from 2010-2019.
+We proceed to examine the trends in all-cause death rates per 1000 population from 2010 to 2021, encompassing both COVID-related deaths and non-COVID-related fatalities, as well as the trend in excess death rates per 100 population in 2020 and 2021 in the subsequent section.
+@fig-all-cause-death-rate presents the trends in all-cause death rates per 1000 population from 2010 to 2021 for India, South Korea, Russia, and the United States, utilizing data from @tbl-cleaned-death-rates.
+From 2010 to 2019, a general gradual decrease in death rates is evident for Russia and India, juxtaposed with a gradual increase observed in South Korea and the United States.
+The dashed line represents the line of best fit for the years 2010-2019.
 
@@
 
-We see here in @fig-all-cause-death-rate that there is a sudden increase in the death rate in 2020 and 2021, and the increase in Russia's death rate is much higher than that of the other countries.
-The essential point to note here is that in the case of India, which has a GHS index lower than that of Russia, the rising trend observed in 2020 and 2021 was still not as excessive as in the case of Russia.
-The dotted line represents the line of best fit for the trend from 2010-2021.
-As we go from comparing Russia to South Korea, we see that in the case of South Korea, both the dashed (2010-2019) and the dotted (2010-2021) are very close together implying that the country did not experience an unanticipated number of deaths during COVID.
-However, in the case of Russia, we see the two lines diverging implying that there were unforeseen deaths during that period.
+From @fig-all-cause-death-rate, a notable surge in death rates is observed in 2020 and 2021, particularly pronounced in Russia compared to other countries.
+It is noteworthy that despite India's lower GHS index compared to Russia, the observed increase in death rates during 2020 and 2021 was not as substantial as in Russia.
+The dotted line represents the trend line for the years 2010-2021.
+Comparing Russia to South Korea, the dashed (2010-2019) and dotted (2010-2021) lines closely align for South Korea, indicating a relatively consistent trend without a significant increase in unexpected deaths during COVID-19.
+Conversely, for Russia, the divergence between the two lines implies a notable increase in unforeseen deaths during this period.
 
 ## Excess Death Rates {#sec-results-excess-rates}
-This observation leads us to @fig-excess-death-rate, where we present the excess death rate for the four countries for 2020 and 2021.
-As mentioned above, the excess death rates were calculated by extrapolating the dashed line of best-fit from years 2010-2019 in @fig-all-cause-death-rate representing the trend in all cause death rates to 2020 and 2021.
-This way we were able to find the expected death rate for each of the four countries and subsequently we were able to plot the difference between the expected death rate and the actual death rate called the excess death rate for 2020 and 2021 for all four countries in @fig-excess-death-rate.
+Considering the likelihood of undercounting COVID-19 deaths, our focus shifted to analyzing excess death rates for the years 2020 and 2021 to provide a comprehensive measure of the pandemic's overall mortality impact.
+Drawing inspiration from Nuzzo and Ledesma's methodology (@citeRepPaper), which compared observed deaths during the pandemic to expected deaths extrapolated from historical trends, we utilized a similar model.
+Extrapolating the dashed line of best-fit from 2010-2019 to 2020 and 2021 allowed us to determine the expected death rate for each country.
+The subsequent plotting of the difference between expected and actual death rates, termed the excess death rate, unveiled insights into the pandemic's impact on mortality.
+
+Our calculated values of excess death rates are presented in the figure labeled "Excess Death Rate" (@fig-excess-death-rate), showcasing the rates for Russia, India, South Korea, and the United States in 2020 and 2021.
+Simultaneously, the figure labeled "All-Cause Death Rate" (@fig-all-cause-death-rate) illustrates the trend in all-cause death rates for the same period.
+This simple models provides a nuanced understanding of the excess death rates, contributing valuable insights into the broader mortality implications of the COVID-19 pandemic.
 
@@
 
-We see in @fig-excess-death-rate the massive difference between the excess death rates for Russia and the other countries.
-Comparing Russia to South Korea we see that, as seen in @fig-all-cause-death-rate, South Korea was very faintly affected by the COVID-19 pandemic, however Russia saw a sudden rise in deaths from 2019 to 2021.
-This shows us that the observations here are analogous to the plot in @fig-all-cause-death-rate.
-When we compare Russia with India, which as aforementioned was deemed a less secure country in the case of epidemics, we see that the excess death rates for Russia in both 2020 and 2021 are much higher than those for India.
-This observation will further aid us in making meaningful discussions in @sec-discussion.
+In @fig-excess-death-rate, a considerable contrast is evident, particularly highlighting the substantial difference in excess death rates for Russia compared to other countries.
+The comparison with South Korea, as depicted in @fig-all-cause-death-rate, reinforces the minimal impact of the COVID-19 pandemic on South Korea, while Russia experienced a notable surge in deaths from 2019 to 2021.
+Comparing Russia to India, designated as less secure in the context of epidemics, the excess death rates for Russia in both 2020 and 2021 significantly exceed those for India.
+This observation will serve as a crucial point of reference in our ensuing discussions in @sec-discussion.
 
@@
 
-Finally, we will now explore the relationship between the excess death rate seen above and the GHS index for a subset of countries per 100,000 population.
-@fig-excess-death-rate-ghs shows us a subset of countries scattered around the plot according to their excess death rate and the GHS index shown in the Appendix @tbl-appendix.
-The green line shows the general decreasing trend which implies that as the GHS index increases, the excess death rate tends to decrease, i.e.
-the excess death rate for more "secure" or well-prepared countries is generally lower.
+Finally, we delve into the relationship between the excess death rate and the Global Health Security (GHS) index for a subset of countries per 100,000 population.
+@fig-excess-death-rate-ghs visually represents this subset of countries scattered on the plot based on their excess death rate and GHS index, as detailed in Appendix @tbl-appendix.
+The green line signifies a general decreasing trend, suggesting that as the GHS index increases, the excess death rate tends to decrease – indicating that more "secure" or well-prepared countries generally experience lower excess death rates.
 
-Russia, labelled in red, is seen to be significantly higher than the line of best fit, which implies that Russia's excess death rate was much worse than it should have been.
-We observe that countries like Brazil and Indonesia, which have approximately the same GHS index as Russia, are nearer to the line of best fit than Russia.
-India, although having a lower GHS index than Russia is seen to have a much lower than general excess death rate.
+Russia, highlighted in red, stands notably above the line of best fit, indicating that its excess death rate exceeded expectations.
+In contrast, countries like Brazil and Indonesia, with comparable GHS indices to Russia, align more closely with the line of best fit.
+India, despite having a lower GHS index than Russia, exhibits a substantially lower excess death rate.
 
@@
 
-The implications from this figure will be further studied in @sec-discussion.
+@fig-excess-death-rate-ghs prompts further exploration in @sec-discussion, where we will delve into the implications and potential factors contributing to Russia's disproportionately higher excess death rate compared to countries with similar or even lower GHS indices.
 

### 2024-02-14 scripts/03-test_data.R

@@ -1,6 +1,6 @@
 #### Preamble ####
 # Purpose: Tests for the cleaned dataset to check if the requirements match that of the simulated data.
-# Author: Aryaman Sharma
+# Author: Aryaman Sharma, Aviral Bhardwaj, Janel Gilani
 # Date: 13 February 2024
 # Contact: aryaman.sharma@mail.utoronto.ca
 # License: MIT

### 2024-02-15 paper/paper.qmd

@@
 ---
-title: "Evaluating Russia's Preparedness and Response to the COVID-19 Pandemic: An Analysis of Excess Death Rates and the Global Health Security Index"
+title: "Evaluating Russia's Response to the COVID-19 Pandemic: An Analysis of High Excess Death Rates Despite High GHS Index"
 author: 
@@
 
-The global impact of the COVID-19 pandemic on public health, foreign policy, and economies has been profound, resulting in an unprecedented loss of life worldwide over the past five years (World Health Organization [@citeWHOeffect]).
-With over 774 million confirmed cases and more than 7 million reported deaths, these figures, considered to be undercounted, suggest an actual death toll ranging from 18.2 to 33.5 million based on excess mortality estimates (WikiCOVID [@citeWikiCOVID]).
+The global impact of the COVID-19 pandemic on public health, foreign policy, and economies has been profound, resulting in an unprecedented loss of life worldwide over the past five years [@citeWHOeffect].
+With over 774 million confirmed cases and more than 7 million reported deaths, these figures, considered to be undercounted, suggest an actual death toll ranging from 18.2 to 33.5 million based on excess mortality estimates [@citeWikiCOVID].
 The staggering number of deaths has prompted a critical examination of countries presumed to be well-prepared to handle such outbreaks and the factors contributing to their extensive impact.
@@
 Among the most severely affected nations, including India, the United States of America, Brazil and Mexico, Russia, a nation deemed well-prepared according to various metrics, was specifically chosen for this study due to its unexpected and disproportionately high death toll in the face of the pandemic.
-With over 230,000 reported COVID-19-related deaths, Russia holds the highest mortality rate in Europe and the second-highest in Asia (ThinkGlobalHealth [@citeThinkGlobalHealth]; Carnegie Russia [@citeCarnegieRussia]).
+With over 230,000 reported COVID-19-related deaths, Russia holds the highest mortality rate in Europe and the second-highest in Asia [@citeThinkGlobalHealth]; [@citeCarnegieRussia].
 This selection aims to uncover unique insights into the challenges faced by ostensibly prepared nations.
@@
 Nuzzo and Jorge R.
-Ledesma in their study, utilizing the Global Health Security (GHS) Index as an indicator of a country's preparedness against epidemics or pandemics like COVID-19 (RepPaper [@citeRepPaper]).
+Ledesma in their study, utilizing the Global Health Security (GHS) Index as an indicator of a country's preparedness against epidemics or pandemics like COVID-19 [@citeRepPaper].
 The choice of GHS is deliberate, considering its comprehensive evaluation of various health security capacities.
@@
 The structure of this paper comprises three sections: Data, Results, and Discussion.
-In the Data section (@sec-data), we discuss the data source and the process of measuring and cleaning the datasets from the replicated paper (RepPaper [@citeRepPaper]).
+In the Data section (@sec-data), we discuss the data source and the process of measuring and cleaning the datasets from the replicated paper [@citeRepPaper].
 The Results section (@sec-results) summarizes the findings, presenting maps and plots to illustrate trends.
@@
 Considering the likelihood of undercounting COVID-19 deaths, our focus shifted to analyzing excess death rates for the years 2020 and 2021 to provide a comprehensive measure of the pandemic's overall mortality impact.
-Drawing inspiration from Nuzzo and Ledesma's methodology (@citeRepPaper), which compared observed deaths during the pandemic to expected deaths extrapolated from historical trends, we utilized a similar model.
+Drawing inspiration from Nuzzo and Ledesma's methodology [@citeRepPaper], which compared observed deaths during the pandemic to expected deaths extrapolated from historical trends, we utilized a similar model.
 Extrapolating the dashed line of best-fit from 2010-2019 to 2020 and 2021 allowed us to determine the expected death rate for each country.
@@
 
-## Findings: Russia, Excess Deaths and what went wrong during the COVID-19 pandemic {#sec-first-point}
+## Findings: Russia, Excess Deaths, and Challenges during the COVID-19 Pandemic {#sec-first-point}
 
-In this paper, we have replicated the paper by Nuzzo and Ledesma [@citeRepPaper], where they focused their analysis on the United States and studying how the world's best prepared country still couldn't control the pandemic.
-Our paper has instead shifted the focus to another global power and fairly well prepared country Russia, where we found out that Russia, when compared to a subset of other countries:(1) had one of the highest increase in all cause death rate in 2020, and (2) had one of the highest excess death rate caused by the pandemic in 2020 and 2021, as observed in @fig-all-cause-death-rate and @fig-excess-death-rate.
-Our paper also found and observed the relationship between the GHS index of a country and its excess death rate in @fig-excess-death-rate-ghs, where Russia was seen as an outlier, where it lied a lot higher than the line of best fit implying a higher excess death rate than the other countries around it, even with a similar GHS index.
+This study, inspired by Nuzzo and Ledesma's work [@citeRepPaper], which scrutinized the United States' pandemic response, redirects the focus towards Russia, a globally influential and reasonably well-prepared nation.
+Our examination revealed that, in comparison to a subset of other countries, Russia experienced (1) a substantial increase in the all-cause death rate in 2020 and (2) one of the highest excess death rates attributed to the pandemic in 2020 and 2021, as illustrated in @fig-all-cause-death-rate and @fig-excess-death-rate.
+The exploration of the relationship between a country's Global Health Security (GHS) index and its excess death rate in @fig-excess-death-rate-ghs identified Russia as an outlier, significantly surpassing the line of best fit, indicating a higher excess death rate relative to countries with similar GHS indices.
 
-The main question that this paper tackles is, what exactly did Russia do wrong during the pandemic considering it was the first country to roll out a vaccine?
-According to a report by Carnegie Endowment for International Peace, by late October 2021, only one-third of the Russian Population had been fully vaccinated against the virus, where other countries like Argentina and China had 56% and 76% of their populations respectively, fully vaccinated by then [@citeCarnegieRussia].
-According to reports, one of the main concerns for people being skeptical about the Sputnik V vaccine was that they did not trust that the vaccine would be effective or even safe to use.
-Another reason for the residents of Russia being skeptical was that Russian President Putin hesitated for months after the release of the vaccine to himself get vaccinated [@citeCarnegieRussia].
-I believe that this must have slowed down the rate at which people in Russia were getting vaccinated hence leading to unprecedented deaths as a result.
+The primary inquiry arising from this paper pertains to understanding the shortcomings in Russia's pandemic response, especially considering its early deployment of a vaccine.
+Reports from the Carnegie Endowment for International Peace indicate that, by late October 2021, only one-third of the Russian population had received full vaccination, trailing behind countries like Argentina and China with 56% and 76% vaccination coverage, respectively [@citeCarnegieRussia].
+Vaccine skepticism among the Russian populace, fueled by concerns about efficacy and safety, contributed to hesitancy, exacerbated by President Putin's delay in personally receiving the Sputnik V vaccine [@citeCarnegieRussia].
+This hesitancy likely impeded the vaccination rate, potentially leading to the unprecedented death toll.
 
-Another possible reason for Russia not being able to tackle the pandemic well was that the quarantine measures during the outbreak in 2020 and 2021 were generally very sporadic and short [@citeNewYorker].
-It was reported that at a time when the cases testing positive reached nearly ten thousand a day, the government put lockdown restrictions but that too only for a week, before everything opened back up.
-I believe this would have been a huge contributing factor to the extremely high excess deaths as more people would have contracted the virus coming in contact with each, hence spreading it more.
-Additionally, keeping in mind that at this time, as aforementioned, Russia only had approximately one thirds of its population vaccinated, this would have hit the residents of the country really hard.
+Another plausible factor contributing to Russia's pandemic challenges was the implementation of sporadic and short-lived quarantine measures in 2020 and 2021, as reported by [@citeNewYorker].
+Lockdown restrictions, initiated when daily positive cases reached nearly ten thousand, were only enforced for a brief week before being lifted.
+This lax approach to containment measures may have facilitated widespread transmission, particularly given the low vaccination coverage at that time, amplifying the impact on the population.
 
-In @sec-second-point, we will discuss the measures that Russia could have taken to tackle the pandemic in a better way, and also measures that can be taken for future pandemics.
+In @sec-second-point, we will delve into potential strategies Russia could have employed for a more effective pandemic response and discuss broader measures for addressing future pandemics.
 
@@
 
-I believe that, as fast as Russia was at releasing the vaccine, maybe the government could have focused on mass producing and advertising the vaccine as a necessity as well.
-I believe that as soon as there is any skepticism in regards with vaccines, people will choose not to get one.
-As mentioned above, this will lead to the virus spreading even faster and eventually to more deaths too.
-In order to prevent this, the government should show the residents of the country that the vaccine is trustworthy and effective.
-It is essential to note that in order for this to work, the vaccine's efficacy and safety will also have to be proven to prevent from any ethical implications rising.
+While Russia demonstrated swiftness in developing a vaccine, it could have augmented its efforts in mass production and promotion, emphasizing the vaccine's necessity.
+Addressing vaccine skepticism is crucial, as public perception significantly influences vaccination rates.
+To foster trust, the government must transparently demonstrate the vaccine's efficacy and safety, mitigating ethical concerns and bolstering public confidence.
 
-It was claimed that at the beginning of the outbreak, Russia was trying to downplay the threat of virus, however, with time they realized how lethal it actually was [@citeCarnegieRussia].
-According to me, it is very essential for a country to recognize the potential threat from any epidemic or outbreak and then make laws, and rules and regulations around it to quickly contain the situation before it gets worse.
-For countries as large as Russia, it might be a particularly challenging task to get such a large population that is spread out under control, but if the right procedures are set in place to tackle such situations, it can be manageable.
-Additionally, it is essential in these situations to also have firm rules in place for quarantining and lockdowns to prevent the virus from spreading.
-This is especially even more important in the early stages of the outbreak when there is no vaccine in place to prevent the citizens of a country from contracting the virus.
-Some countries during the COVID-19 pandemic were very strict with their restrictions and lockdowns, and hence experienced way lesser deaths when compared to other countries which were more relaxed.
-In the future too, for Russia to be able to control the situation early on, better and more strict regulations will be required.
+Initial attempts to downplay the virus's severity may have hindered the country's response, as acknowledged later [@citeCarnegieRussia].
+Recognizing the gravity of an epidemic is pivotal, prompting proactive legislation and regulations to swiftly contain outbreaks.
+Despite the logistical challenges posed by Russia's vast population, effective measures and protocols can be established to manage such situations.
+Robust laws and regulations, coupled with stringent quarantine and lockdown protocols, are imperative, particularly in the absence of a vaccine during the early stages of an outbreak.
+Countries that implemented strict measures during the COVID-19 pandemic experienced lower mortality rates compared to those with more lenient approaches.
+For Russia to proactively manage future outbreaks, the establishment of enhanced and stringent regulations is paramount.
 
@@
 
-It is essential to take into account an unintended but highly likely bias that might have taken place in this paper.
-While working on the paper, we realized that in creating @fig-all-cause-death-rate, @fig-excess-death-rate and @fig-excess-death-rate-ghs, selection bias might have been involved in make decisions on which countries to plot on the graphs.
-Specifically, we used 4 countries in @fig-all-cause-death-rate and @fig-excess-death-rate to show the unique trends for Russia during the pandemic.
-Maybe if we chose a set of different countries, some countries would have performed more like Russia and consequently would have impacted our analysis.
-Similarly, in @fig-excess-death-rate-ghs choosing a different subset of countries might generate a completely different line of best fit which might not have satisfied our exploration in the paper.
-However, at the same time, it is essential to note that the set of countries chosen in this paper has mostly been random and some countries like India and the United States have been shown constantly to create parallels between the impact of the pandemic on both the countries.
+Recognizing the potential for unintended biases in our analysis is imperative.
+In constructing @fig-all-cause-death-rate, @fig-excess-death-rate, and @fig-excess-death-rate-ghs, there is a possibility of selection bias influencing our decision on which countries to include.
+The choice of countries, including Russia, in these figures might impact the observed trends.
+A different set of countries could yield alternative patterns and potentially alter our conclusions.
+However, it is crucial to note that the selection was mostly random, with consistent inclusion of countries like India and the United States to draw meaningful parallels between the pandemic's impact on various nations.
 
-It should be noted that different countries employ varying methods of reporting their data, including mortality rates and excess death rates.
-Therefore, this study primarily relied on observational data, but it's crucial to account for differential biases in the data.
-For instance, some countries might underreport their deaths, while others might overreport them.
-Such discrepancies directly impact the analysis conducted in this paper.
+It is essential to acknowledge the inherent variability in how countries report data, including mortality and excess death rates.
+This study heavily relies on observational data, necessitating an understanding of potential biases.
+Variations in reporting practices, such as underreporting or overreporting deaths, introduce nuances that directly influence the analysis.
 
-Furthermore, COVID-19 policies were predominantly implemented at the state and local levels, introducing variability throughout the pandemic.
-Differences in policy implementation among these communities, coupled with the presence of unmeasured differences, complicate the ability to draw causal inferences.
+Moreover, the implementation of COVID-19 policies occurred predominantly at the state and local levels, introducing additional complexity and potential unmeasured differences.
+The diverse approaches among communities make drawing definitive causal inferences challenging.
 
-There are a few ethical implications that we found while working on the paper.
-Mainly, sometimes the data can be incomplete or hidden which is misleading while analyzing that data.
-It is essential to keep that into account when looking at a report as, if for in this case any data for the deaths were incomplete, that would directly impact how we deem Russia to have tackled the pandemic.
-This is not just applicable for Russian data, but if other countries' data was missing, it would alter a major part of the analysis done above.
+Ethical considerations also come into play, particularly concerning incomplete or concealed data.
+Incomplete information can be misleading, impacting the evaluation of a country's pandemic response.
+This ethical concern extends beyond Russia, encompassing any country with missing or incomplete data, highlighting the need for transparency and accuracy in reporting to ensure robust and unbiased analyses.
 
@@
 
-While working on the paper we came across a few weaknesses that can be built on to make the report even more sound.
-Firstly, we could have tried to take into account the selection bias by choosing the countries except Russia randomly.
-It is obviously impractical to include all 195 countries in all plots, hence choosing a completely random subset of countries could have reduced the selection bias.
-However, it is also essential to note that it is preferable to compare Russia with countries with the same infrastructure and stature in the world, as that makes the comparisons more sound.
-Comparing Russian data with random countries might have led to this paper being less informative and less analytical.
-Then, another weakness was that while making @fig-map it was observed that the raw data file did not actually contain the GHS indices for all the countries around the world, for example, specifically Greenland was excluded from the dataset, which was the reason why some of the countries in the figure are shaded in grey showing that data was not available for them.
-If all the data was present for the figure, the analysis could have been even more sound.
+During the course of this paper, several weaknesses were identified that could be addressed to enhance its robustness.
+One notable limitation is the potential for selection bias in country inclusion for analysis.
+While attempting to mitigate this bias, selecting a random subset of countries excluding Russia could have reduced its impact.
+However, it's important to note that comparisons with countries of similar infrastructure and global stature provide more meaningful insights.
+Random selection might have compromised the analytical depth and informativeness of the paper.
 
-As a further scope, an extension to this report could include taking into account the proportion of population vaccinated for a country in 2021.
-Taking this factor into account, we would be able to clearly point out one of the prominent reasons why Russia might have lacked in their attempt to tackle the pandemic.
-Alternatively, we can flip the question for this paper around and delve deeper into the countries with the lowest GHS indices which surprisingly did better than expected.
-This would help us analyze the paper in positive notion, taking into account all the right things they did that led to them tackling the outbreak better than many well-prepared countries.
+Another weakness pertains to the availability of data.
+In the construction of @fig-map, it was evident that the dataset lacked GHS indices for certain countries, such as Greenland, leading to incomplete shading in the figure.
+A comprehensive dataset would have strengthened the analysis and provided more nuanced insights into global health security preparedness.
 
-In conclusion, we have observed that over 2020 and 2021, Russia failed to properly control the spread of the pandemic, and also effectively get its population vaccinated, which consequently lead to a large number of positive cases and deaths around the country.
-For future epidemics and outbreaks, Russia needs to make strict regulations and procedures which will help them contain the virus and not let it spread in the country and cause enormous amount of deaths.
+As a future scope, incorporating vaccination rates into the analysis for 2021 could offer valuable insights into pandemic control efforts.
+Examining the proportion of vaccinated population across countries would help elucidate factors contributing to Russia's challenges in containing the pandemic.
+Additionally, flipping the focus to countries with lower GHS indices that managed the outbreak effectively could provide lessons in effective pandemic management.
+Understanding the strategies and interventions that facilitated success in these countries could inform future preparedness and response efforts.
+
+Regarding the selection of countries for comparison, our criteria primarily focused on countries with diverse demographics, varied GHS indices, and significant global influence.
+While biases may exist, the inclusion of countries like the United States and India aimed to provide a comprehensive analysis of pandemic responses across different contexts.
+However, future research could explore comparisons with countries of smaller populations and lower GDPs to further enrich the analysis and mitigate potential biases.
+
+In conclusion, the analysis underscores Russia's shortcomings in controlling the COVID-19 pandemic, particularly in vaccination efforts and containment measures.
+Moving forward, implementing stringent regulations and proactive strategies are imperative to prevent widespread transmission and mitigate the impact of future epidemics and outbreaks.
 

### 2024-02-15 paper/paper.qmd
Minor edit to appendix plot.

@@
 #| tbl-cap: "Cleaned Data showing Excess Death Rate and GHS index for a subset of countries"
-#| tbl-cap-location: bottom
 #| warning: false
